---
title: "MFSR Water Sr EDA"
subtitle: "Exploratory summary of compiled water ⁸⁷Sr/⁸⁶Sr results"
format: html
editor: visual
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(here)
library(sf)

dat <- read_csv(here("results/compiled_sr_data.csv"), show_col_types = FALSE)
```

This doc reads the output of `R/01_compile_sr_data.R` (`results/compiled_sr_data.csv`). Re-run that script first if you've added new samples or a new ICP-MS batch.

## Data-quality checks

Samples with no coordinates (should be empty -- everything should resolve through `master_site_lookup.csv`):

```{r}
#| label: qc-missing-coords
dat |>
  filter(is.na(lat_dd) | is.na(long_dd)) |>
  select(sample_id, site_id, waterbody_name)
```

Samples with no Sr result yet (expected for very recent collections still waiting on lab turnaround; anything old showing up here is worth chasing down):

```{r}
#| label: qc-missing-sr
dat |>
  filter(is.na(sr_87)) |>
  select(sample_id, site_id, waterbody_name, collection_date)
```

Duplicate `sample_id` values (should be empty -- the compile script already collapses analytical re-runs, so any duplicates here indicate a real sample-log data-entry problem):

```{r}
#| label: qc-dup-ids
dat |>
  count(sample_id) |>
  filter(n > 1)
```

## Samples per stream and year

```{r}
#| label: samples-per-stream-year
dat |>
  mutate(year = year(collection_date)) |>
  count(waterbody_name, year) |>
  pivot_wider(names_from = year, values_from = n, values_fill = 0) |>
  arrange(waterbody_name)
```

## Sr summary by stream

```{r}
#| label: sr-summary-by-stream
dat |>
  filter(!is.na(sr_87)) |>
  summarise(
    n = n(),
    sr_min = min(sr_87),
    sr_mean = mean(sr_87),
    sr_max = max(sr_87),
    .by = waterbody_name
  ) |>
  arrange(desc(sr_mean))
```

## Sr by waterbody

```{r}
#| label: fig-sr-by-waterbody
#| fig-width: 8
#| fig-height: 7
dat |>
  filter(!is.na(sr_87)) |>
  ggplot(aes(x = reorder(waterbody_name, sr_87), y = sr_87)) +
  geom_point(size = 3) +
  coord_flip() +
  theme_bw(base_size = 14) +
  labs(y = expression(''^87*'Sr/'^86*'Sr'), x = "")
```

## Map of sampled sites

Mean Sr value per site, plotted against the MFSR stream network.

```{r}
#| label: fig-site-map
#| message: false
#| fig-width: 8
#| fig-height: 8

flowline <- st_read(here("data/spatial/flowline_pn17_nsi_mfsr.gpkg"), quiet = TRUE)

site_means <- dat |>
  filter(!is.na(sr_87)) |>
  filter(waterbody_name != "Wilson Creek") |>
  summarise(sr_87 = mean(sr_87), lat_dd = first(lat_dd), long_dd = first(long_dd), .by = site_id) |>
  st_as_sf(coords = c("long_dd", "lat_dd"), crs = 4326) |>
  st_transform(st_crs(flowline))

ggplot() +
  geom_sf(data = flowline, color = "steelblue", linewidth = 0.3) +
  geom_sf(data = site_means, aes(color = sr_87), size = 2.5) +
  scale_color_viridis_c(name = expression(''^87*'Sr/'^86*'Sr')) +
  theme_minimal(base_size = 12) +
  labs(title = "Mean site Sr isotope ratio")
```

For interactive panning/zooming during fieldwork planning, `mapview::mapview(site_means, zcol = "sr_87")` is a quick alternative to the static map above.
