---
title: "2026 Tributary Sampling Plan"
subtitle: "Geology- and GIS-informed site selection for this 2026 MFSR spawning tributaries"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(sf)
library(here)
library(knitr)

litho_colors <- c(
  "Water"                                         = "#3E7CB1",
  "Igneous, intrusive"                            = "#E8763F",
  "Igneous, volcanic"                             = "#A83C2E",
  "Igneous and Metamorphic, undifferentiated"     = "#4FA89B",
  "Metamorphic, undifferentiated"                 = "#7CB86B",
  "Metamorphic, sedimentary clastic"              = "#C9B54A",
  "Metamorphic and Sedimentary, undifferentiated" = "#B98CC4",
  "Sedimentary, clastic"                          = "#D9A441",
  "Sedimentary, undifferentiated"                 = "#C2B8A3",
  "Unconsolidated, undifferentiated"              = "#F2E9A8"
)


fl <- st_read(here("data/spatial/flowline_pn17_nsi_mfsr.gpkg"),
  quiet = TRUE
)
lith <- st_read(here("data/spatial/geology/idaho_lithology.shp"),
  quiet = TRUE
) |>
  st_make_valid() |>
  st_transform(st_crs(fl))
hu8 <- st_read(here("data/spatial/hu8_mfsr.gpkg"),
  quiet = TRUE
) |>
  st_transform(st_crs(fl))
sites <- read_csv(here("data/master_site_lookup.csv"),
  show_col_types = FALSE
)
hamann <- st_read(here("data/spatial/IsoWorld.gdb"),
  layer = "Hamaan2012_SR_WaterSamples",
  quiet = TRUE
) |>
  st_transform(4326)
```

This doc answers one question per target tributary: **given what we know about MFSR bedrock geology, where along this stream should a new site go to best capture real Sr variation?**

**Method**: each named tributary's NHD flowline reaches (`data/spatial/flowline_pn17_nsi_mfsr.gpkg`) are ordered upstream-to-downstream using drainage area (`TotDASqKM`, a standard proxy for position along a single named stream absent full network topology) and matched to the nearest bedrock polygon in the USGS State Geologic Map Compilation (`data/spatial/geology/idaho_lithology.shp`). This gives, for every tributary: (a) what fraction of its length drains each rock unit, and (b) where along its length it crosses from one unit into another — a **lithologic contact**, and the single best predictor of where Sr will shift most sharply. Candidate new sites are placed near contacts closest to the 1/3 and 2/3 marks of each tributary's length (so 2-3 new sites end up reasonably spread out — not evenly spaced by default, since an evenly spaced site could easily land in the middle of one uniform unit and miss the contrast entirely.

**Caveat up front**: this identifies the geologically *ideal* spot. It says nothing about whether that exact point is reachable on foot, whether it's a maintained trail, or what the current stream/trail conditions are. Treat every coordinate below as a target to aim near, not a GPS waypoint to hit exactly; adjust to the nearest safe, accessible point.

```{r}
#| label: analysis-functions
#| code-summary: "Show the tributary analysis functions"

# define drainage names and codes -----
drainage_defs <- list(
  list(
    drainage = "Sulphur Creek",
    names = c("Sulphur Creek", "North Fork Sulphur Creek"), 
    code = "SUL"),
  list(
    drainage = "Pistol Creek", 
    names = c("Pistol Creek", "Little Pistol Creek"),
    code = "PIS"),
  list(
    drainage = "Rapid River",
    names = c("Rapid River"), 
    code = "RAP"),
  list(
    drainage = "Loon Creek",
    names = c("Loon Creek"),
    code = "LOO"),
  list(
    drainage = "Warm Spring Creek", 
    names = c("Warm Spring Creek", "South Fork Warm Spring Creek"), 
    code = "WSC"),
  list(
    drainage = "Mayfield Creek", 
    names = c("Mayfield Creek", "East Fork Mayfield Creek", "West Fork Mayfield Creek"), 
    code = "MAY"),
  list(
    drainage = "Indian Creek",
    names = c("Indian Creek"),
    code = "IND"),
  list(
    drainage = "Camas Creek", 
    names = c("Camas Creek"),
    code = "CAM"),
  list(
    drainage = "West Fork Camas Creek",
    names = c("West Fork Camas Creek", "South Fork Camas Creek"),
    code = "WFC"),
  list(
    drainage = "Yellowjacket Creek", 
    names = c("Yellowjacket Creek", "West Fork Yellowjacket Creek"),
    code = "YJC"),
  list(
    drainage = "Wilson Creek",
    names = c("Wilson Creek"),
    code = "WIL"),
  list(
    drainage = "Big Creek",
    names = c("Big Creek"),
    code = "BIG",
    partner = "Colden Baxter (U Idaho) and students"),
  list(
    drainage = "Rush Creek",
    names = c("Rush Creek"),
    code = "RSH"),
  list(
    drainage = "Beaver Creek (Big Ck drainage)",
    names = c("Beaver Creek"), 
    code = "BVB",
    lat_min = 45, 
    partner = "Payette NF (possible)"),
  list(
    drainage = "Monumental Creek",
    names = c("Monumental Creek", "North Fork Monumental Creek", "West Fork Monumental Creek"), 
    code = "MON",
    partner = "Payette NF (upper reach)"),
  list(
    drainage = "Marble Creek",
    names = c("Marble Creek"), 
    code = "MRB", 
    partner = "Payette NF (possible, if they go in)")
)

# analysis functions ----
analyze_trib <- function(def) {
  seg <- fl |> filter(GNIS_NAME %in% def$names)
  if (!is.null(def$lat_min)) {
    ll <- st_transform(seg, 4326)
    keep <- st_coordinates(st_centroid(st_geometry(ll)))[, 2] >= def$lat_min
    seg <- seg[keep, ]
  }
  if (nrow(seg) == 0) return(NULL)
  seg <- seg |> arrange(TotDASqKM)
  pts <- suppressWarnings(st_centroid(seg))
  joined <- st_join(pts, lith["GENERALIZE"], join = st_nearest_feature)
  seg$GENERALIZE <- joined$GENERALIZE
  seg$cumlen <- cumsum(seg$LENGTHKM)
  ll_pts <- st_transform(pts, 4326)
  seg$lat <- st_coordinates(ll_pts)[, 2]
  seg$lon <- st_coordinates(ll_pts)[, 1]

  total_len <- sum(seg$LENGTHKM)
  comp <- st_drop_geometry(seg) |>
    summarise(len = sum(LENGTHKM), .by = GENERALIZE) |>
    mutate(pct = round(100 * len / total_len, 1)) |>
    arrange(desc(pct))

  d <- st_drop_geometry(seg)
  changes <- which(d$GENERALIZE[-1] != d$GENERALIZE[-nrow(d)])
  crossings <- tibble()
  if (length(changes) > 0) {
    crossings <- map_dfr(changes, function(i) {
      tibble(
        from_unit = d$GENERALIZE[i], to_unit = d$GENERALIZE[i + 1],
        cumlen_km = round((d$cumlen[i] + d$cumlen[i + 1]) / 2, 1),
        lat = round((d$lat[i] + d$lat[i + 1]) / 2, 5),
        lon = round((d$lon[i] + d$lon[i + 1]) / 2, 5)
      )
    })
  }

  list(
    drainage = def$drainage, code = def$code, partner = def$partner,
    total_len_km = round(total_len, 1), composition = comp, crossings = crossings
  )
}

pick_candidates <- function(r, fracs = c(0.33, 0.66)) {
  if (is.null(r) || nrow(r$crossings) == 0) return(tibble())
  total_len <- r$total_len_km
  tol <- max(4, total_len * 0.15)
  map_dfr(fracs, function(f) {
    target <- total_len * (1 - f)
    idx <- which.min(abs(r$crossings$cumlen_km - target))
    best <- r$crossings[idx, ]
    if (abs(best$cumlen_km - target) <= tol) {
      tibble(
        dist_upstream_km = round(total_len - best$cumlen_km, 1),
        lat = best$lat, lon = best$lon,
        rationale = paste0(best$from_unit, " ↔ ", best$to_unit)
      )
    } else {
      tibble(
        dist_upstream_km = round(total_len * f, 1), lat = NA, lon = NA,
        rationale = "no major contact nearby — place by even spacing / access"
      )
    }
  }) |> distinct(lat, lon, .keep_all = TRUE)
}

results <- map(drainage_defs, analyze_trib)
names(results) <- map_chr(drainage_defs, "drainage")

report_drainage <- function(def) {
  r <- results[[def$drainage]]
  if (is.null(r)) {
    cat("No flowline match found for", def$drainage, "-- check GNIS_NAME spelling.\n\n")
    return(invisible())
  }
  cand <- pick_candidates(r)
  cat(sprintf("**%s (`%s`)** -- %.1f km of mapped channel.\n\n", def$drainage, def$code, r$total_len_km))
  print(kable(r$composition |> rename(Lithology = GENERALIZE, `km` = len, `% of length` = pct)))
  cat("\n")
  if (nrow(cand) > 0) {
    cat("Candidate new site positions (distance upstream of the `-010` confluence site):\n\n")
    print(kable(cand |> rename(`km upstream` = dist_upstream_km, Rationale = rationale)))
  } else {
    cat("No lithologic contact detected along the mapped channel -- geologically uniform; ",
        "space new sites evenly based on access rather than geology.\n", sep = "")
  }
  cat("\n")
}
```

## Overview map

```{r}
#| label: fig-overview-map
#| fig-width: 9
#| fig-height: 9
#| message: false

lith_clip <- st_intersection(lith, st_union(hu8))
existing_sites_sf <- sites |>
  st_as_sf(coords = c("long_dd", "lat_dd"), crs = 4326) |>
  st_transform(st_crs(fl))

ggplot() +
  geom_sf(data = lith_clip, aes(fill = GENERALIZE), color = NA, alpha = 0.6) +
  geom_sf(data = fl, color = "grey20", linewidth = 0.15) +
  geom_sf(data = hamann, color = "red", size = 1, shape = 4) +
  geom_sf(data = existing_sites_sf, color = "black", size = 1.5) +
  scale_fill_manual(values = litho_colors, name = "Bedrock unit") + 
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", legend.text = element_text(size = 7)) +
  guides(fill = guide_legend(nrow = 4)) +
  labs(
    title = "MFSR basin: bedrock lithology and sample sites",
    caption = "Black dots = existing sample sites (master_site_lookup.csv). Red x's = Hamann & Kennedy (2012) Big Creek water samples."
  )
```

```{r}
#| label: fig-map-tribs
#| fig-width: 9
#| fig-height: 9
#| message: false

ggplot() +
  geom_sf(data = lith_clip, aes(fill = GENERALIZE), color = NA, alpha = 0.6) +
  geom_sf(data = fl, color = "grey20", linewidth = 0.15) +
  geom_sf(data = fl |> filter(GNIS_NAME=="Middle Fork Salmon River"), 
          color = "blue", linewidth = 1) +
  geom_sf(
    data = fl |> filter(GNIS_NAME %in% c(
      "Sulphur Creek", 
      "Rapid River", 
      "Pistol Creek", 
      "Indian Creek", 
      "Marble Creek", 
      "Loon Creek", 
      "Warm Spring Creek",
      "Mayfield Creek",
      "East Fork Mayfield Creek",
      "West Fork Mayfield Creek",
      "Camas Creek", 
      "Yellowjacket Creek",
      "West Fork Camas Creek",
      "Bear Valley Creek", 
      "Elk Creek",
      "Bearskin Creek",
      "Marsh Creek", 
      "Knapp Creek",
      "Cape Horn Creek",
      "Wilson Creek",
      "Big Creek", 
      "Monumental Creek", 
      "Rush Creek", 
      "Beaver Creek", 
      )), 
    color = "blue", linewidth = 0.4) +
  geom_sf(data = existing_sites_sf, color = "black", size = 1.5) +
  scale_fill_manual(values = litho_colors, name = "Bedrock unit") + 
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", legend.text = element_text(size = 7)) +
  guides(fill = guide_legend(nrow = 4)) +
  labs(title = "MFSR basin: bedrock lithology, sample sites, and major tribs")
```

Bedrock is dominated by two igneous provinces of very different age (Cretaceous Idaho Batholith granite and Eocene Challis Volcanics/intrusives), with a real Precambrian metamorphic component in the northeast part of the basin (Yellowjacket, Big Creek headwaters area) — see `docs/review_biogeochem_tracers.qmd` for the full area-weighted breakdown and why this predicts sharp, mappable Sr contrasts in exactly those tributaries that cross from one province into another.

## Partner-covered sites — not part of Bryan's own itinerary

Colden Baxter (University of Idaho) and students will collect Big Creek while out with students this summer. Payette National Forest biologists will help with upper Monumental Creek, and possibly Marble Creek if they go in. These three drainages are analyzed below for completeness (the geology doesn't stop mattering just because someone else is collecting), but the actual collection logistics/instructions for those collaborators belong in `field/partner_sample_collection_guide.qmd`, not here. Rush Creek and (Big Creek's) Beaver Creek are **not** currently covered by a partner as far as this plan knows — confirm with Baxter's crew whether their Big Creek trip extends to those sub-tributaries before assuming they're covered.

## Sulphur, Pistol, Rapid, Loon, Indian corridor

These five drainages sit along the Bear Valley/Marsh-to-Boundary Creek stretch of the MFSR and are the ones the Thurow redd-survey memo (`docs/memo_thurow_redd_sampling_alternatives.docx`) describes real foot-travel logistics for: a documented pattern is to shuttle to Pistol Creek's headwaters, survey/hike down Pistol (and Little Pistol) to the mainstem, then hike up Indian Creek to a shuttle waiting at *its* headwaters. That same shape (hike down one tributary, up the neighboring one, shuttle at both ends) is a natural fit for placing new Sr sites along the way, since you're walking the whole drainage rather than just visiting the mouth.

```{r}
#| label: fig-sulphur
#| results: asis
report_drainage(drainage_defs[[1]])
```

```{r}
#| label: fig-pistol
#| results: asis
report_drainage(drainage_defs[[2]])
```

Pistol Creek's channel alternates repeatedly between granite and valley-fill (Quaternary alluvium) rather than crossing a single clean contact — expected given it's flowing through a broad batholith-floored valley, not between two different bedrock provinces. The two candidate positions above are still reasonable anchor points, but don't expect as sharp a contrast here as in a drainage that crosses between actual different-age bedrock (e.g. Indian Creek, below).

```{r}
#| label: fig-rapid
#| results: asis
report_drainage(drainage_defs[[3]])
```

Rapid River is 100% Idaho Batholith granite along its entire mapped length — geologically uniform. There's no lithologic reason to place a new site at any particular spot; if you add one beyond the existing `RAP-010`, place it based on access (trail conditions, camp spacing) rather than geology.

```{r}
#| label: fig-loon
#| results: asis
report_drainage(drainage_defs[[4]])
```

```{r}
#| label: fig-indian
#| results: asis
report_drainage(drainage_defs[[7]])
```

Indian Creek is the most geologically interesting of this group: it's roughly half Challis volcanics, half Idaho Batholith granite, with the contact sitting well inside the drainage rather than at either end — a real candidate for the single sharpest Sr contrast among all the tributaries sampled this way (see also the Warm Spring/Mayfield discussion below, which shows the same volcanic/granite juxtaposition even more repeatedly).

## Loon Creek's named sub-tributaries: Warm Spring & Mayfield creeks

```{r}
#| label: fig-warmspring
#| results: asis
report_drainage(drainage_defs[[5]])
```

```{r}
#| label: fig-mayfield
#| results: asis
report_drainage(drainage_defs[[6]])
```

Both Warm Spring and Mayfield creeks weave repeatedly between Challis volcanics and Idaho Batholith granite — more evidence that this part of the basin (northwest of Loon Creek's mouth) sits right at the boundary between the two igneous provinces. That's good news for this project (lots of contrast to sample) but means the "candidate" positions above are representative anchor points along a genuinely patchy contact zone, not a single clean line — small adjustments to the nearest accessible spot won't meaningfully change what you'd measure.

## Camas Creek and its named sub-tributaries: WF Camas & Yellowjacket creeks

```{r}
#| label: fig-camas
#| results: asis
report_drainage(drainage_defs[[8]])
```

```{r}
#| label: fig-wfcamas
#| results: asis
report_drainage(drainage_defs[[9]])
```

```{r}
#| label: fig-yellowjacket
#| results: asis
report_drainage(drainage_defs[[10]])
```

Yellowjacket Creek stands out here: it's the only drainage in this year's target list with a substantial share (\~46%) of Precambrian Belt Supergroup metamorphic rock alongside Idaho Batholith granite. Per `docs/review_biogeochem_tracers.qmd`, ancient metamorphic terrain tends to be a highly radiogenic "anchor" — this is likely to be one of the most distinctive tributary signatures in the whole project, and worth prioritizing if time runs short.

## Wilson Creek

```{r}
#| label: fig-wilson
#| results: asis
report_drainage(drainage_defs[[11]])
```

Wilson Creek is short but geologically mixed (granite, metamorphic, and valley-fill all present) — a compact drainage where 2 well-placed sites should still capture most of the real variation.

## Big Creek drainage (partner-covered) and its sub-tributaries

```{r}
#| label: fig-big
#| results: asis
report_drainage(drainage_defs[[12]])
```

```{r}
#| label: fig-rush
#| results: asis
report_drainage(drainage_defs[[13]])
```

```{r}
#| label: fig-beaverbig
#| results: asis
report_drainage(drainage_defs[[14]])
```

```{r}
#| label: fig-monumental
#| results: asis
report_drainage(drainage_defs[[15]])
```

```{r}
#| label: fig-marble
#| results: asis
report_drainage(drainage_defs[[16]])
```

Big Creek itself is the most geologically diverse single tributary in this whole plan (five different rock categories along its length, including a meaningful metamorphic/sedimentary component) — this is exactly the kind of drainage where Baxter's crew collecting good spatial coverage along the mainstem will be valuable. Rush Creek and Beaver Creek (this Big Creek one, `BVB` — not to be confused with Marsh Creek's `BVR`, see `data/site_id_convention.md`) both show real igneous/metamorphic contrasts too and are worth confirming coverage on with Baxter's team.

## Punch list

```{r}
#| label: tbl-punchlist
punchlist <- map_dfr(drainage_defs, function(def) {
  r <- results[[def$drainage]]
  tibble(
    drainage = def$drainage,
    code = def$code,
    mapped_km = if (!is.null(r)) r$total_len_km else NA,
    dominant_unit = if (!is.null(r)) r$composition$GENERALIZE[1] else NA,
    n_contacts = if (!is.null(r)) nrow(r$crossings) else NA,
    partner = if (!is.null(def$partner)) def$partner else "Bryan (own itinerary)"
  )
})
kable(punchlist)
```

Priorities if the season runs short on time: **Yellowjacket Creek** (only metamorphic-heavy drainage in Bryan's own list) and **Indian Creek** (cleanest single volcanic/granite contact) give the most new isotopic information per site visited. **Rapid River** is the lowest-priority stop for a *new* upstream site specifically, since it's geologically uniform — the existing `RAP-010` already characterizes it well.
