---
title: "2026 Tributary Sampling Plan"
subtitle: "Geology- and GIS-informed site selection for this 2026 MFSR spawning tributaries"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(sf)
library(here)
library(knitr)
library(ggspatial)

# scale bar + north arrow, added to every map on this doc
map_furniture <- function() {
  list(
    annotation_scale(location = "bl", width_hint = 0.25, text_cex = 0.7),
    annotation_north_arrow(location = "tr", which_north = "true", height = unit(0.9, "cm"), width = unit(0.9, "cm"))
  )
}

## Colors match John's manuscript rock-type scheme (5 categories). Our raw
## USGS SGMC `GENERALIZE` field has 9 categories plus "Water" (river-channel
## digitizing artifact -- no lakes in the MFSR, so it's dropped entirely).
## Everything below groups to these 5 for both maps and tables.
litho_colors <- c(
  "Plutonic"            = "hotpink",
  "Volcanic"             = "saddlebrown",
  "Metasedimentary"      = "darkolivegreen4",
  "Sedimentary"          = "steelblue1",
  "Quaternary sediments" = "yellow"
)

lith_group_map <- c(
  "Igneous, intrusive"                             = "Plutonic",
  "Igneous, volcanic"                              = "Volcanic",
  "Igneous and Metamorphic, undifferentiated"      = "Metasedimentary",
  "Metamorphic, undifferentiated"                  = "Metasedimentary",
  "Metamorphic, sedimentary clastic"               = "Metasedimentary",
  "Metamorphic and Sedimentary, undifferentiated"  = "Metasedimentary",
  "Sedimentary, clastic"                           = "Sedimentary",
  "Sedimentary, undifferentiated"                  = "Sedimentary",
  "Unconsolidated, undifferentiated"               = "Quaternary sediments"
)

fl <- st_read(here("data/spatial/flowline_pn17_nsi_mfsr.gpkg"), quiet = TRUE)
lith <- st_read(here("data/spatial/geology/idaho_lithology.shp"), quiet = TRUE) |>
  st_make_valid() |>
  st_transform(st_crs(fl)) |>
  filter(GENERALIZE != "Water") |>
  mutate(lith_group = recode(GENERALIZE, !!!lith_group_map))
hu8 <- st_read(here("data/spatial/hu8_mfsr_dissolved.gpkg"), quiet = TRUE) |> st_transform(st_crs(fl))
sites <- read_csv(here("data/master_site_lookup.csv"), show_col_types = FALSE)
hamann <- st_read(here("data/spatial/IsoWorld.gdb"), layer = "Hamaan2012_SR_WaterSamples", quiet = TRUE) |>
  st_transform(4326)

# roads/trails for field-logistics reference on the zoomed maps
roads <- bind_rows(
  st_read(here("data/spatial/MFS_RoadsTrails_Albers96/Idaho_Roads_Clip5000_Albers96.shp"), quiet = TRUE) |> st_transform(st_crs(fl)) |> select(geometry),
  st_read(here("data/spatial/MFS_RoadsTrails_Albers96/USFS_Roads_Clip5000_Albers96.shp"), quiet = TRUE) |> st_transform(st_crs(fl)) |> select(geometry)
)
trails <- st_read(here("data/spatial/MFS_RoadsTrails_Albers96/USFS_Trails_Clip5000_Albers96.shp"), quiet = TRUE) |> st_transform(st_crs(fl))
```

Colors follow the same rock-family scheme John uses in his manuscript figures (Plutonic/Volcanic/Metasedimentary/Sedimentary/Quaternary sediments). 

This doc answers one question per target tributary: **given what we know about MFSR bedrock geology, where along this stream should a new site go to best capture real Sr variation?**

**Method**: each named tributary's NHD flowline reaches (`data/spatial/flowline_pn17_nsi_mfsr.gpkg`) are ordered upstream-to-downstream using drainage area (`TotDASqKM`, a standard proxy for position along a single named stream absent full network topology) and matched to the nearest bedrock polygon in the USGS State Geologic Map Compilation (`data/spatial/geology/idaho_lithology.shp`). This gives, for every tributary: (a) what fraction of its length drains each rock unit, and (b) where along its length it crosses from one unit into another — a **lithologic contact**, and the single best predictor of where Sr will shift most sharply. Candidate new sites are placed near contacts closest to the 1/3 and 2/3 marks of each tributary's length (so 2-3 new sites end up reasonably spread out — not evenly spaced by default, since an evenly spaced site could easily land in the middle of one uniform unit and miss the contrast entirely.

::: {.callout-note collapse="true"}
## Tributary identification reference (GNIS_ID)

Several stream names in this basin are shared by more than one physical channel, so every tributary below is selected by verified `GNIS_ID` rather than name string. `GNIS_NAME` is kept alongside for readability.

| Drainage | `GNIS_ID` used | Other stream with the same name (not used here) |
|---|---|---|
| Sulphur Creek | 391601 | 391603, an unrelated stream elsewhere in the basin |
| Warm Spring Creek | 400163 | 400162, a different "Warm Spring Creek" flowing directly into the MFSR |
| Indian Creek | 383652 | 383645, a small unrelated stream that flows into Loon Creek |
| Rush Creek | 389731 | 400063, an unrelated stream near Loon Creek |
| Beaver Creek (Big Ck) | 377253 | 377241, Marsh Creek's Beaver Creek (`BVR`) |
| Loon Creek | 399100 | -- extends ~23 km beyond the named reaches down to the true confluence |
:::

**Caveat up front**: this identifies the geologically *ideal* spot. It says nothing about whether that exact point is reachable on foot, whether it's a maintained trail, or what the current stream/trail conditions are. Treat every coordinate below as a target to aim near, not a GPS waypoint to hit exactly; adjust to the nearest safe, accessible point.

**Sample upstream or downstream of a contact?** Downstream, as a default. Water carries a rock unit's dissolved Sr signature only after it's had a chance to weather that rock, so a sample taken just *upstream* of a contact still reflects the *previous* (upstream) unit -- it hasn't touched the new one yet. Sampling just *downstream* lets the water pick up the new unit's contribution first. In practice "just downstream" should mean a bit past the contact rather than immediately at it, since water needs a short distance to mix across the channel width; a few hundred meters to a kilometer downstream is a reasonable rule of thumb where access allows. See `docs/review_biogeochem_tracers.qmd` for the underlying Sr geochemistry.

```{r}
#| label: analysis-functions
#| code-summary: "Show the tributary analysis functions"

# define drainages by verified GNIS_ID (see QA/QC callout above for how these
# were resolved). The FIRST gnis_id listed is the "trunk" used for lithology
# composition/contact-crossing/candidate-site analysis; any additional ids
# are named side-forks (e.g. Little Pistol Creek, East/West Fork Mayfield)
# that join the trunk as separate parallel channels -- shown on the map for
# context only, never merged into the trunk's up-to-down sequence (see the
# QA/QC callout on forks below for why that merge silently breaks). Where a
# tributary here is itself a named sub-tributary of ANOTHER stream in this
# list, parent_gnis_id/parent_name draws that parent stream on the zoomed map
# too, per Bryan's request to see e.g. Mayfield/Warm Spring alongside Loon.
drainage_defs <- list(
  list(
    drainage = "Sulphur Creek", 
    gnis_ids = c("391601", "387799"), 
    code = "SUL"),
  list(
    drainage = "Pistol Creek", 
    gnis_ids = c("388566", "384918"), 
    code = "PIS"),
  list(
    drainage = "Rapid River", 
    gnis_ids = c("389107"), 
    code = "RAP"),
  list(
    drainage = "Loon Creek", 
    gnis_ids = c("399100"),
    code = "LOO", 
    fill_gap = TRUE),
  list(
    drainage = "Warm Spring Creek", 
    gnis_ids = c("400163", "394337"),
    code = "WSC", 
    parent_gnis_id = "399100", 
    parent_name = "Loon Creek"),
  list(
    drainage = "Mayfield Creek", 
    gnis_ids = c("386556", "381130", "393142"), 
    code = "MAY", 
    parent_gnis_id = "399100", 
    parent_name = "Loon Creek"),
  list(
    drainage = "Indian Creek",
    gnis_ids = c("383652"), 
    code = "IND"),
  list(
    drainage = "Camas Creek",
    gnis_ids = c("378735"), 
    code = "CAM"),
  list(
    drainage = "West Fork Camas Creek",
    gnis_ids = c("393066"), 
    code = "WFC",
    parent_gnis_id = "378735",
    parent_name = "Camas Creek"),
  list(
    drainage = "Yellowjacket Creek",
    gnis_ids = c("393757", "393228"), 
    code = "YJC", 
    parent_gnis_id = "378735",
    parent_name = "Camas Creek"),
  list(
    drainage = "Wilson Creek",
    gnis_ids = c("393547"), 
    code = "WIL"),
  list(
    drainage = "Big Creek", 
    gnis_ids = c("377483"), 
    code = "BIG",
    partner = "Colden Baxter (U Idaho) and students"),
  list(
    drainage = "Rush Creek",
    gnis_ids = c("389731"), 
    code = "RSH", 
    parent_gnis_id = "377483",
    parent_name = "Big Creek"),
  list(
    drainage = "Beaver Creek (Big Ck drainage)", 
    gnis_ids = c("377253"), 
    code = "BVB", 
    partner = "Payette NF (possible)",
    parent_gnis_id = "377483", 
    parent_name = "Big Creek"),
  list(
    drainage = "Monumental Creek",
    gnis_ids = c("387122", "387739", "393151"), 
    code = "MON", 
    partner = "Payette NF (upper reach)", 
    parent_gnis_id = "377483", 
    parent_name = "Big Creek"),
  list(
    drainage = "Marble Creek", 
    gnis_ids = c("386455"), 
    code = "MRB", 
    partner = "Payette NF (unconfirmed -- Bryan likely collects if they don't go in)", 
    parent_gnis_id = "377483",
    parent_name = "Big Creek"),
  # headwaters reach: Big Hole (Marsh Ck + Bear Valley Ck confluence, the
  # official start of the MFSR) down to the Boundary Creek put-in -- Bryan
  # already has upper Marsh/Bear Valley samples but nothing in these two
  # currently-unsampled stretches. Restricted to a TotDASqKM window on an
  # already-resolved single-GNIS_ID stream, so no gap-fill needed.
  list(
    drainage = "Lower Marsh Creek (MRS-010 to Big Hole)", 
    gnis_ids = c("386480"), 
    code = "MRS", 
    totda_range = c(360, 386)),
  list(
    drainage = "MFSR headwaters (Big Hole to Boundary Creek)", 
    gnis_ids = c("386841"), 
    code = "MFSR", 
    totda_range = c(884, 1022))
)

# directed downstream walk to absorb unnamed "gap" reaches (e.g. Loon Creek's
# missing lower third) without wandering onto side tributaries: at each step,
# only the touching reach with the LARGEST TotDASqKM is the true trunk
# continuation, and we stop as soon as that reach turns out to be a real,
# differently-named waterbody (the actual confluence). IMPORTANT: only
# Loon Creek actually has a confirmed attribute gap -- applying this to every
# tributary is itself a bug we hit (Warm Spring Creek's downstream walk from
# its own mouth immediately entered Loon Creek's unnamed gap and kept
# absorbing Loon Creek's own reaches, since fill_downstream_gap has no way to
# know it had already crossed into a different stream's territory once both
# are unnamed). So this is now opt-in per drainage (`fill_gap = TRUE`), set
# only where a real gap was confirmed against a known `-010` site -----
fill_downstream_gap <- function(seed_ids, fl, max_iter = 60) {
  named <- fl |> filter(GNIS_ID %in% seed_ids)
  current <- named[which.max(named$TotDASqKM), ]
  pool <- fl |> filter(COMID != current$COMID[1])
  filled_ids <- c()
  for (i in seq_len(max_iter)) {
    touching <- suppressMessages(st_filter(pool, current, .predicate = st_is_within_distance, dist = 1))
    touching <- touching |> filter(TotDASqKM > current$TotDASqKM[1])
    if (nrow(touching) == 0) break
    nxt <- touching[which.max(touching$TotDASqKM), ]
    if (!is.na(nxt$GNIS_NAME)) break # reached a real, different named waterbody -- stop
    filled_ids <- c(filled_ids, nxt$COMID)
    pool <- pool |> filter(COMID != nxt$COMID)
    current <- nxt
  }
  bind_rows(named, fl[fl$COMID %in% filled_ids, ])
}

# analysis functions ----

# fetch reaches for a set of GNIS_IDs, honoring totda_range/fill_gap the same
# way regardless of whether it's the trunk or a fork
fetch_reaches <- function(ids, def) {
  if (length(ids) == 0) return(fl[0, ])
  if (!is.null(def$totda_range)) {
    fl |> filter(GNIS_ID %in% ids, TotDASqKM >= def$totda_range[1], TotDASqKM <= def$totda_range[2])
  } else if (isTRUE(def$fill_gap)) {
    fill_downstream_gap(ids, fl)
  } else {
    fl |> filter(GNIS_ID %in% ids)
  }
}

# order by TotDASqKM, join lithology, compute cumulative distance -- using
# st_point_on_surface (guaranteed to sit ON the line) rather than st_centroid
# (which for a bent/curvy reach can land off the line entirely -- the direct
# cause of candidate markers not landing on the stream line)
add_lith_and_order <- function(x) {
  if (nrow(x) == 0) return(x)
  x <- x |> arrange(TotDASqKM)
  pts <- suppressWarnings(st_point_on_surface(x))
  x$lith_group <- st_join(pts, lith["lith_group"], join = st_nearest_feature)$lith_group
  x$cumlen <- cumsum(x$LENGTHKM)
  ll <- st_transform(pts, 4326)
  x$lat <- st_coordinates(ll)[, 2]
  x$lon <- st_coordinates(ll)[, 1]
  x
}

analyze_trib <- function(def) {
  primary_id <- def$gnis_ids[1]
  fork_ids <- setdiff(def$gnis_ids, primary_id)

  # trunk only: composition, contact-crossings, and candidate sites all come
  # from this single ordered line. Named side-forks (fork_ids) are fetched
  # separately and never merged into the trunk's TotDASqKM sequence -- doing
  # that was the bug behind off-line candidates on Pistol Creek (vs. Little
  # Pistol Creek) and Mayfield Creek (vs. its East/West Forks): forks are
  # parallel channels whose drainage areas interleave with the trunk's rather
  # than continuing it, so sorting them together produces "contacts" between
  # geometrically unrelated points on two different lines.
  seg <- fetch_reaches(primary_id, def)
  seg_forks <- fetch_reaches(fork_ids, def)
  if (nrow(seg) == 0) return(NULL)

  seg <- add_lith_and_order(seg)
  seg_forks <- add_lith_and_order(seg_forks)

  total_len <- sum(seg$LENGTHKM)
  comp <- st_drop_geometry(seg) |>
    summarise(len = sum(LENGTHKM), .by = lith_group) |>
    mutate(pct = round(100 * len / total_len, 1)) |>
    arrange(desc(pct))

  d <- st_drop_geometry(seg)
  changes <- which(d$lith_group[-1] != d$lith_group[-nrow(d)])
  crossings <- tibble()
  if (length(changes) > 0) {
    crossings <- map_dfr(changes, function(i) {
      tibble(
        from_unit = d$lith_group[i], to_unit = d$lith_group[i + 1],
        cumlen_km = round((d$cumlen[i] + d$cumlen[i + 1]) / 2, 1),
        lat = round((d$lat[i] + d$lat[i + 1]) / 2, 5),
        lon = round((d$lon[i] + d$lon[i + 1]) / 2, 5)
      )
    })
  }

  list(
    drainage = def$drainage, code = def$code, partner = def$partner,
    total_len_km = round(total_len, 1), composition = comp, crossings = crossings,
    seg = seg, seg_forks = seg_forks, fork_ids = fork_ids,
    parent_gnis_id = def$parent_gnis_id, parent_name = def$parent_name
  )
}

pick_candidates <- function(r, fracs = c(0.33, 0.66)) {
  if (is.null(r) || nrow(r$crossings) == 0) return(tibble())
  total_len <- r$total_len_km
  tol <- max(4, total_len * 0.15)
  map_dfr(fracs, function(f) {
    target <- total_len * (1 - f)
    idx <- which.min(abs(r$crossings$cumlen_km - target))
    best <- r$crossings[idx, ]
    if (abs(best$cumlen_km - target) <= tol) {
      tibble(
        dist_upstream_km = round(total_len - best$cumlen_km, 1),
        lat = best$lat, lon = best$lon,
        rationale = paste0(best$from_unit, " <-> ", best$to_unit),
        source = "Analysis"
      )
    } else {
      tibble(
        dist_upstream_km = round(total_len * f, 1), lat = NA, lon = NA,
        rationale = "no major contact nearby - place by even spacing / access",
        source = "Analysis"
      )
    }
  }) |> distinct(lat, lon, .keep_all = TRUE)
}

# unlike pick_candidates() (which selects 2-3 well-spaced NEW site spots),
# this surfaces EVERY detected contact -- used for the Bear Valley/Marsh
# Creek retrospective review below, where the question is "how do existing
# sites compare to all real contacts," not "where should new sites go"
all_crossings <- function(r) {
  if (is.null(r) || nrow(r$crossings) == 0) return(tibble())
  r$crossings |>
    transmute(
      dist_upstream_km = round(r$total_len_km - cumlen_km, 1),
      lat, lon,
      rationale = paste0(from_unit, " <-> ", to_unit),
      source = "Analysis"
    )
}

# combine the automatic contact-based picks with any hand-picked sites Bryan
# adds as the season firms up -- both kept side by side (tagged by `source`)
# rather than one replacing the other, unless noted otherwise for a specific
# drainage (e.g. Warm Spring Creek's two close-together auto picks were
# consolidated into one at Bryan's request)
manual_site <- function(dist_upstream_km, lat, lon, rationale) {
  tibble(dist_upstream_km = dist_upstream_km, lat = lat, lon = lon, rationale = rationale, source = "Manual")
}

results <- map(drainage_defs, analyze_trib)
names(results) <- map_chr(drainage_defs, "drainage")

# Bear Valley Creek and upper Marsh Creek (above MRS-010) are a retrospective
# check on 2024 site placement, not 2026 targets -- kept OUT of drainage_defs
# on purpose so they don't show up in the punch list or the "all candidate
# sites this season" map, but still get the same analyze_trib()/report_
# drainage()/map_drainage() treatment by appending directly into `results`
bvc_def <- list(drainage = "Bear Valley Creek", gnis_ids = c("377200"), code = "BVC")
mrs_upper_def <- list(drainage = "Marsh Creek (upper, above MRS-010)", gnis_ids = c("386480"), code = "MRS", totda_range = c(0, 360))
results[[bvc_def$drainage]] <- analyze_trib(bvc_def)
results[[mrs_upper_def$drainage]] <- analyze_trib(mrs_upper_def)

# manual overrides for cases where the generic 1/3-2/3-contact heuristic
# isn't the right call (see the headwaters section) -- keyed by drainage name
# so both report_drainage() and map_drainage() (and the final summary map)
# all read the same recommended site rather than the map showing one thing
# and the table showing another
manual_overrides <- list(
  "Lower Marsh Creek (MRS-010 to Big Hole)" = manual_site(
    4.5, 44.427, -115.194,
    "Metasedimentary <-> Plutonic ('halfway' point lands almost exactly on this contact)"
    ),
  "MFSR headwaters (Big Hole to Boundary Creek)" = manual_site(
    16.2, 44.45431, -115.2304,
    "just below Big Hole -- post-confluence mixed signature; nudged onto a trail per OnX field review"
    ),
  # field-verified via OnX (2026-07-10): all of these were auto-picked or
  # coarsely-placed and landed upslope off the actual channel, or somewhere
  # with no reasonable access -- coordinates below are Bryan's corrected,
  # trail-adjacent picks, still in the same lithologic unit/contact zone
  "Sulphur Creek" = bind_rows(
    manual_site(4.4, 44.54112, -115.34786, "Quaternary sediments <-> Plutonic contact -- moved onto the channel/trail"),
    manual_site(12.5, 44.54337, -115.39404, "~5 km downstream of the North Fork Sulphur Creek confluence -- moved onto the channel/trail")
  ),
  "Pistol Creek" = bind_rows(
    manual_site(4.1, 44.72169, -115.19505, "just below the Little Pistol Creek confluence"),
    manual_site(13.5, 44.70957, -115.29234, "Quaternary sediments <-> Plutonic contact"),
    manual_site(19.9, 44.68706, -115.35482, "Plutonic <-> Quaternary sediments contact"),
    manual_site(NA, 44.72483, -115.20771, "near the mouth of Little Pistol Creek itself -- simplified from the original ~11 km-up contact pick to an easier, more accessible spot right at the confluence")
  ),
  "Rapid River" = manual_site(20.3, 44.57013, -115.07105, "trailhead with road access"),
  # expanded from 3 to 5 sites per Bryan's review of the spawning-location
  # dataset -- lower two now bracket the narrow metasedimentary inlier's
  # contacts directly, upper site pushed further upstream for better coverage
  "Indian Creek" = bind_rows(
    manual_site(0.3, 44.77525, -115.1219, "near-mouth spatial coverage"),
    manual_site(4.6, 44.79249, -115.1326, "lower Volcanic <-> Plutonic contact"),
    manual_site(6.3, 44.79784, -115.13715, "narrow metasedimentary inlier, downstream edge"),
    manual_site(6.3, 44.80273, -115.15513, "narrow metasedimentary inlier, upstream edge"),
    manual_site(15.5, 44.83038, -115.22374, "upper-drainage Plutonic <-> Volcanic contact -- better upper coverage")
  ),
  # expanded from 3 to 4 sites -- the two lower contacts become three
  # (one new, per Bryan), trailhead site unchanged
  "Marble Creek" = bind_rows(
    manual_site(7.8, 44.78388, -114.97391, "second contact downstream of the upper Volcanic <-> Plutonic candidate"),
    manual_site(9.9, 44.80634, -114.97713, "added for spatial coverage between the two bracketing contacts"),
    manual_site(11.0, 44.81619, -114.98009, "Volcanic <-> Plutonic contact"),
    manual_site(34.2, 44.95702, -115.09996, "top of drainage by trailhead")
  ),
  # expanded from 3 to 5 sites -- lower contact and both sedimentary-patch
  # edges moved onto trail-accessible spots, plus two new mid-drainage sites
  "Loon Creek" = bind_rows(
    manual_site(15.8, 44.69593, -114.76016, "Volcanic <-> Plutonic contact"),
    manual_site(25.0, 44.6354, -114.74627, "mid-drainage spatial coverage"),
    manual_site(32.1, 44.60136, -114.80704, "additional coverage approaching the sedimentary patch"),
    manual_site(34.9, 44.57025, -114.84637, "sedimentary patch, downstream edge"),
    manual_site(40.4, 44.54133, -114.85956, "sedimentary patch, upstream edge")
  ),
  # replaces (not adds to) the two auto-picked candidates, which bracketed the
  # same small granite inlier only ~1.6 km apart -- consolidated to one
  # in-between site plus one near the Loon Creek confluence, per Bryan
  "Warm Spring Creek" = bind_rows(
    manual_site(14.2, 44.62334, -114.59868, "midpoint of the granite inlier (averages the two closely-spaced auto picks) -- moved onto trail"),
    manual_site(1.5, 44.65062, -114.7171, "just upstream of the confluence with Loon Creek")
  ),
  "Mayfield Creek" = bind_rows(
    manual_site(4.2, 44.53957, -114.80626, "just below the East/West Fork confluence, on the short main-stem trunk"),
    manual_site(NA, 44.50394, -114.74393, "on West Fork Mayfield Creek, just upstream of its volcanic/granite contact -- moved a bit further upstream (from the contact itself) per Russ, to sit solidly in the volcanic unit rather than right on the boundary; the trunk and East Fork sites already cover the plutonic side"),
    manual_site(NA, 44.54327, -114.77872, "on East Fork Mayfield Creek, moved further along the fork for trail access -- Russ wants to hike up the East Fork a bit, so worth a stop while he's up there")
  ),
  "Monumental Creek" = bind_rows(
    manual_site(24.5, 44.98076, -115.16526, "near the Monumental/Marble trailhead, upper drainage"),
    manual_site(13.5, 45.08099, -115.12124, "Volcanic <-> Metasedimentary contact -- moved onto trail per OnX review"),
    manual_site(1.9, 45.13964, -115.15895, "confirmed Metasedimentary, ~1.9 miles (3.1 km) upstream of the confluence -- a real hike in from the confluence site below, but squarely in the unit rather than straddling the contact"),
    manual_site(0.0, 45.16149, -115.1303, "at Hamann & Kennedy's (2012) 'Big Cr. upstream of Monumental Cr.' sample point, ~1.5 km up Big Creek from the actual Monumental Creek mouth -- Plutonic, and the closest real reference site to the confluence; replaces both the previous confluence-mouth estimate and the old 'site F' pick (~8 km upstream, Metasedimentary, not Plutonic)")
  ),
  # Hamann & Kennedy (2012) sites snap almost exactly (0 m) onto our resolved
  # trunk line -- sites D and C aren't added here since they're already
  # closely matched by the existing 20.5/45 km auto candidates (see prose
  # below); E, A, and B fill in gaps the auto analysis didn't pick
  "Big Creek" = bind_rows(
    manual_site(19.4, 45.12953, -114.95042, "Volcanic <-> Quaternary sediments contact -- moved onto trail per OnX review"),
    manual_site(45.9, 45.17451, -115.18171, "Metasedimentary <-> Plutonic contact -- moved onto trail per OnX review"),
    manual_site(8.7, 45.10261, -114.84607, "Hamann & Kennedy (2012) 'Lower Big Cr.' (site E)"),
    manual_site(60.9, 45.10630, -115.32839, "Hamann & Kennedy (2012) 'Upper Big Cr.' (site A)"),
    manual_site(58.2, 45.12779, -115.31390, "Hamann & Kennedy (2012) 'Big Cr. downstream of Logan Cr.' (site B)"),
    manual_site(30.9, 45.14177, -115.05557, "mid-drainage spatial coverage within the volcanic stretch between the 19.4 km and 45.9 km contact-based candidates")
  ),
  # expanded from 3 to 4 sites -- lower spatial-coverage site and the lower
  # contact moved onto trail-accessible spots, plus one new mid-drainage site
  "Camas Creek" = bind_rows(
    manual_site(6.3, 44.89367, -114.65237, "between CAM-010 and the Yellowjacket Creek confluence (~6.4 km) -- lithologically uniform here (Plutonic), so this is for spatial coverage; per Russ, worth grabbing given how much variation Camas and Yellowjacket show overall"),
    manual_site(14.1, 44.87871, -114.57145, "Volcanic <-> Plutonic contact"),
    manual_site(23.1, 44.82205, -114.49718, "added for mid-drainage spatial coverage"),
    manual_site(37.0, 44.73435, -114.49296, "Plutonic <-> Volcanic contact")
  ),
  "Rush Creek" = bind_rows(
    pick_candidates(results[["Rush Creek"]]),
    manual_site(0.0, 45.08826, -114.88196, "Hamann & Kennedy (2012) 'Rush Cr.' reference site -- near the Big Creek confluence and reachable via the ranch/trail network; priority stop since everything upstream is remote with no road")
  ),
  # upstream contact-based candidates dropped -- no spawning up there, so no
  # reason to ask Baxter's crew to detour off the near-confluence stop
  "Beaver Creek (Big Ck drainage)" = manual_site(0.0, 45.17340, -115.24476, "Hamann & Kennedy (2012) 'Beaver Cr.' reference site -- near the Big Creek confluence; priority stop, same access logic as Rush Creek"),
  "Yellowjacket Creek" = bind_rows(
    manual_site(7.8, 44.94551, -114.59496, "Metasedimentary <-> Plutonic contact -- moved onto trail per OnX review and spawning-location data"),
    manual_site(15.8, 44.97865, -114.50894, "Plutonic <-> Metasedimentary contact -- moved onto trail per OnX review and spawning-location data")
  ),
  "Wilson Creek" = bind_rows(
    manual_site(7.8, 45.0384, -114.62491, "Quaternary sediments <-> Metasedimentary contact -- moved onto trail per OnX review and spawning-location data"),
    manual_site(9.9, 45.05967, -114.59659, "Plutonic <-> Quaternary sediments contact -- moved onto trail per OnX review and spawning-location data")
  ),
  # retrospective review, not new-site picks -- show every detected contact
  # (all_crossings()) rather than the 2-3-site spacing heuristic
  "Bear Valley Creek" = all_crossings(results[["Bear Valley Creek"]]),
  "Marsh Creek (upper, above MRS-010)" = all_crossings(results[["Marsh Creek (upper, above MRS-010)"]])
)

get_candidates <- function(def) {
  if (!is.null(manual_overrides[[def$drainage]])) return(manual_overrides[[def$drainage]])
  pick_candidates(results[[def$drainage]])
}

# composition/candidate table condensed to one line + one compact table (no
# lat/lon columns -- those are only needed for the map marker, not for
# scanning the text) so each tributary section reads as map -> table -> prose
# rather than map -> table -> table -> prose
report_drainage <- function(def, candidate_label = "Candidate new site positions") {
  r <- results[[def$drainage]]
  if (is.null(r)) {
    cat("No flowline match found for", def$drainage, "-- check GNIS_ID.\n\n")
    return(invisible())
  }
  cand <- get_candidates(def)
  comp_str <- r$composition |>
    mutate(part = sprintf("%.0f%% %s", pct, lith_group)) |>
    pull(part) |>
    paste(collapse = ", ")
  cat(sprintf("**%s (`%s`)** -- %.1f km of mapped channel. Bedrock: %s.\n\n", def$drainage, def$code, r$total_len_km, comp_str))
  if (nrow(cand) > 0) {
    cat(candidate_label, " (distance upstream of the `-010` confluence site):\n\n", sep = "")
    print(kable(cand |> select(source, dist_upstream_km, rationale) |>
      rename(Source = source, `km upstream` = dist_upstream_km, Rationale = rationale)))
  } else {
    cat("No lithologic contact detected along the mapped channel -- geologically uniform; ",
        "space new sites evenly based on access rather than geology.\n", sep = "")
  }
  if (length(r$fork_ids) > 0 && nrow(r$seg_forks) > 0) {
    cat("\nNamed side-fork(s) also shown on the map below (grey lines) as context: ",
        paste(unique(r$seg_forks$GNIS_NAME), collapse = ", "), ".\n", sep = "")
  }
  cat("\n")
}

# zoomed per-tributary map: bedrock context + the (corrected, gap-filled)
# tributary line + existing site(s) + candidate new site(s). Candidates come
# from get_candidates(), which reads manual_overrides where set (e.g. the
# headwaters reach) so the map always matches what report_drainage() prints
# -----
map_drainage <- function(def) {
  r <- results[[def$drainage]]
  if (is.null(r)) return(invisible(NULL))
  seg <- r$seg
  cand <- get_candidates(def)

  # zoom extent covers the trunk AND any named forks, so the confluence with
  # a fork (e.g. Little Pistol Creek, East/West Fork Mayfield) is visible
  zoom_geom <- if (nrow(r$seg_forks) > 0) {
    st_union(st_union(seg), st_union(r$seg_forks))
  } else {
    st_union(seg)
  }
  bbox <- st_bbox(st_buffer(zoom_geom, 3000))
  bbox_poly <- st_as_sfc(bbox)
  lith_zoom <- suppressWarnings(st_crop(lith, bbox))

  # coord_sf's xlim/ylim only set the *view*; when the bbox's aspect ratio
  # doesn't match the figure device, ggplot expands one axis to preserve true
  # geographic aspect, and anything well outside the intended zoom (other
  # sites, the whole basin's stream network) can end up inside that expanded
  # view. Crop every layer to the bbox explicitly -- with nothing left
  # outside the zoom to leak in, and expand = FALSE below to stop coord_sf
  # padding the view further, the map is a true zoom regardless of aspect.
  fl_zoom <- suppressWarnings(st_crop(fl, bbox))
  existing <- sites |>
    filter(stream_code == r$code) |>
    st_as_sf(coords = c("long_dd", "lat_dd"), crs = 4326) |>
    st_transform(st_crs(fl)) |>
    st_filter(bbox_poly, .predicate = st_intersects)
  roads_zoom <- suppressWarnings(st_crop(roads, bbox))
  trails_zoom <- suppressWarnings(st_crop(trails, bbox))
  hamann_zoom <- suppressWarnings(st_crop(st_transform(hamann, st_crs(fl)), bbox))
  hu8_zoom <- suppressWarnings(st_crop(hu8, bbox))

  # white-cased like roads/trails below -- a plain grey50 line at this width
  # nearly vanishes against the dark saddlebrown "Volcanic" fill, which is
  # what made streamlines look absent on zoomed maps over volcanic bedrock
  # (they're still there, just low-contrast; the overview map's lower fill
  # alpha hides the same issue) -----
  p <- ggplot() +
    geom_sf(data = lith_zoom, aes(fill = lith_group), color = NA, alpha = 0.75) +
    geom_sf(data = fl_zoom, color = "white", linewidth = 0.35) +
    geom_sf(data = fl_zoom, color = "grey40", linewidth = 0.15)

  if (nrow(hu8_zoom) > 0) p <- p + geom_sf(data = hu8_zoom, fill = NA, color = "grey30", linewidth = 0.6, linetype = "dashed")

  if (nrow(roads_zoom) > 0) p <- p +
    geom_sf(data = roads_zoom, color = "white", linewidth = 0.7) +
    geom_sf(data = roads_zoom, color = "firebrick", linewidth = 0.35)
  if (nrow(trails_zoom) > 0) p <- p +
    geom_sf(data = trails_zoom, color = "white", linewidth = 0.6) +
    geom_sf(data = trails_zoom, color = "black", linewidth = 0.3, linetype = "22")
  if (nrow(hamann_zoom) > 0) p <- p + geom_sf(data = hamann_zoom, color = "blue", size = 2, shape = 2, stroke = 1)

  # parent stream (e.g. Loon Creek behind Warm Spring/Mayfield), cropped to
  # this same zoom -- typically shows just the confluence-adjacent stretch
  title_suffix <- ""
  if (!is.null(r$parent_gnis_id)) {
    parent_zoom <- suppressWarnings(st_crop(fl |> filter(GNIS_ID == r$parent_gnis_id), bbox))
    if (nrow(parent_zoom) > 0) {
      p <- p + geom_sf(data = parent_zoom, color = "black", linewidth = 1, linetype = "dotted", alpha = 0.7)
      title_suffix <- paste0(" -- tributary of ", r$parent_name)
    }
  }
  # named side-forks (e.g. Little Pistol Creek), shown but not analyzed
  if (nrow(r$seg_forks) > 0) {
    p <- p + geom_sf(data = r$seg_forks, color = "grey35", linewidth = 0.6)
  }

  p <- p +
    geom_sf(data = seg, color = "steelblue4", linewidth = 1.1) +
    scale_fill_manual(values = litho_colors, name = "Bedrock unit", drop = TRUE) +
    map_furniture() +
    coord_sf(xlim = bbox[c("xmin", "xmax")], ylim = bbox[c("ymin", "ymax")], expand = FALSE) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position = "right", legend.text = element_text(size = 7), legend.title = element_text(size = 8),
      plot.caption = element_text(hjust = 0, size = 7)
    ) +
    labs(title = paste0(def$drainage, " (`", def$code, "`)", title_suffix), x = NULL, y = NULL)

  if (nrow(existing) > 0) {
    p <- p +
      geom_sf(data = existing, color = "black", size = 2.5, shape = 17) +
      geom_sf_text(data = existing, aes(label = site_id), size = 2.6, fontface = "bold", nudge_y = 400, check_overlap = TRUE)
  }
  if (nrow(cand) > 0 && any(!is.na(cand$lat))) {
    cand_sf <- cand |>
      filter(!is.na(lat)) |>
      st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
      st_transform(st_crs(fl))
    p <- p + geom_sf(data = cand_sf, color = "red", size = 3, shape = 4, stroke = 1.5)
  }
  p +
    labs(caption = str_wrap(paste0(
      "Black triangle = existing site (labeled). Red x = candidate new site. Grey dashed outline = HUC8 basin edge. ",
      "Red line (white-cased) = road. Black dashed line (white-cased) = trail. Blue triangle = Hamann & Kennedy (2012) reference site.",
      if (nrow(r$seg_forks) > 0) " Grey line = named side-fork (context only)." else "",
      if (title_suffix != "") " Black dotted stream line = parent stream." else ""
    ), width = 100))
}

# saves a copy of a zoomed drainage map (or the all-candidates overview) as a
# static PNG under img/site_maps/ -- the single source of truth for these
# figures, so field/partner_sample_collection_guide.qmd just embeds the image
# instead of re-deriving the same maps with a second copy of this analysis
save_partner_map <- function(p, code) {
  dir.create(here("img/site_maps"), showWarnings = FALSE, recursive = TRUE)
  ggsave(here("img/site_maps", paste0(code, ".png")), p, width = 8, height = 6.5, dpi = 150, bg = "white")
  invisible(p)
}
```

## Overview map

```{r}
#| label: fig-overview-map
#| fig-width: 9
#| fig-height: 9
#| message: false

lith_clip <- st_intersection(lith, st_union(hu8))
existing_sites_sf <- sites |>
  st_as_sf(coords = c("long_dd", "lat_dd"), crs = 4326) |>
  st_transform(st_crs(fl))

# uses the corrected, GNIS_ID-verified + gap-filled tributary geometries
# (results[[...]]$seg), not a raw GNIS_NAME filter -- avoids re-introducing
# the name-collision/gap bugs described in the QA/QC callout above
target_segs <- bind_rows(map(results, function(x) {
  if (is.null(x)) return(NULL)
  bind_rows(x$seg, x$seg_forks)
}))

ggplot() +
  geom_sf(data = lith_clip, aes(fill = lith_group), color = NA, alpha = 0.6) +
  geom_sf(data = hu8, fill = NA, color = "grey30", linewidth = 0.5, linetype = "dashed") +
  geom_sf(data = fl, color = "grey20", linewidth = 0.15) +
  geom_sf(data = fl |> filter(GNIS_NAME == "Middle Fork Salmon River"), color = "blue", linewidth = 1) +
  geom_sf(data = target_segs, color = "blue", linewidth = 0.5) +
  geom_sf(data = hamann, color = "red", size = 1, shape = 4) +
  geom_sf(data = existing_sites_sf, color = "black", size = 1.5) +
  scale_fill_manual(values = litho_colors, name = "Bedrock unit") +
  map_furniture() +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right", legend.text = element_text(size = 7),
    plot.caption = element_text(hjust = 0, size = 7)
  ) +
  guides(fill = guide_legend(nrow = 5)) +
  labs(
    title = "MFSR basin: bedrock lithology and sample sites",
    caption = str_wrap("Black dots = existing sample sites (master_site_lookup.csv). Red x's = Hamann & Kennedy (2012) Big Creek water samples.", width = 100)
  )
```

Bedrock is dominated by two igneous provinces of very different age (Cretaceous Idaho Batholith granite and Eocene Challis Volcanics/intrusives), with a real Precambrian metamorphic component in the northeast part of the basin (Yellowjacket, Big Creek headwaters area) — see `docs/review_biogeochem_tracers.qmd` for the full area-weighted breakdown and why this predicts sharp, mappable Sr contrasts in exactly those tributaries that cross from one province into another.

## Who's covering what — 2026 season plan

Every drainage here is analyzed the same way regardless of who collects it — but worth laying out who's doing what and when, for partner coordination and to sequence Bryan's own remaining trips.

**Colden Baxter (University of Idaho) and students** will collect Big Creek this summer, plus the near-confluence sites on Beaver Creek, Monumental Creek, and Rush Creek (see each drainage's section below for coordinates). Rush Creek beyond its near-confluence spot is remote and difficult, so realistically just one sample there. These coordinates are what go into `field/partner_sample_collection_guide.qmd` for their crew.

**Payette National Forest biologists** will help with upper Monumental Creek, and may also be able to cover upper Beaver Creek (otherwise Bryan). Upper Marble Creek is hopefully also Payette — if that pans out, Bryan may fly into lower Marble Creek to pick up the two lower-drainage sites there instead of hiking the whole thing.

**Russ** will collect Loon Creek, Warm Spring Creek, and Mayfield Creek within the next two weeks (by ~2026-07-23) — see the new manual sites added to those three drainages below, all requested directly by Russ ahead of that trip. He may also cover Camas Creek, likely in September, hiking down the drainage to the river and floating out — which is why a manual site was added between `CAM-010` and the Yellowjacket confluence, worth grabbing given how much Sr variation both Camas and Yellowjacket show.

**Bryan's own remaining itinerary**: Sulphur Creek, Rapid River, possibly extra Marsh/Bear Valley sites, Pistol Creek, Indian Creek, lower Marble Creek, and Yellowjacket Creek. Rough trip logistics:

- **Rapid River** is an easy drive-in, grab-and-leave stop, and could be combined with the Sulphur Creek trip (and possibly extra Marsh/Bear Valley sites) in one outing.
- **Sulphur Creek** is road-accessible but far enough to be a long day trip or a short overnight based out of Boundary Creek campground.
- **Pistol Creek and Indian Creek** call for a 4-5 day trip — in through Pistol, out through Indian, with a shuttle needed.
- **Lower Marble Creek** is a likely fly-in, especially if Payette covers upper Marble.
- **Yellowjacket Creek** should be doable via road access.

## Headwaters

The MFSR officially begins at "Big Hole," where Marsh Creek and Bear Valley Creek meet (confirmed in the flowline data — both mouths sit at the same point). Bryan already has 2024 samples from upper Bear Valley Creek (`BVC-010/020/030`) and upper Marsh Creek (`MRS-010/020/030`); the next two subsections retroactively check those placements against the same contact analysis used elsewhere in this doc. Two stretches remain unsampled: the lower ~8 km of Marsh Creek above Big Hole, and the ~18 km of new MFSR channel from Big Hole down to `MFSR-070` at Boundary Creek.

### Bear Valley Creek (existing sites review)

```{r}
#| label: fig-map-bvc
map_drainage(bvc_def)
```

```{r}
#| label: fig-bvc
#| results: asis
report_drainage(bvc_def, candidate_label = "Detected lithologic contacts (for comparison against existing sites)")
```

Bear Valley Creek is mostly Quaternary valley-fill with a real Plutonic component; the analysis finds four contacts. `BVC-010` and `BVC-030` both sit within a couple km of a real contact — good placement. `BVC-020` sits in a uniform Quaternary stretch with no contact nearby — likely more about spatial/access coverage than lithology. The most-upstream contact falls beyond `BVC-030` entirely — the one gap worth a look if this creek gets revisited.

### Marsh Creek, above MRS-010 (existing sites review)

```{r}
#| label: fig-map-mrsupper
map_drainage(mrs_upper_def)
```

```{r}
#| label: fig-mrsupper
#| results: asis
report_drainage(mrs_upper_def, candidate_label = "Detected lithologic contacts (for comparison against existing sites)")
```

100% Quaternary sediments — no internal contact above `MRS-010`. A useful negative result: `MRS-010/020/030` aren't missing anything here, since the nearest contact (Lower Marsh Creek, below) sits downstream of `MRS-010`. The tight 0.708 cluster these sites show (`eda_water_sr.qmd`) fits a uniform valley-fill unit, not undersampling.

### Lower Marsh Creek

```{r}
#| label: fig-map-lowermarsh
map_drainage(drainage_defs[[17]])
```

```{r}
#| label: fig-lowermarsh
#| results: asis
report_drainage(drainage_defs[[17]])
```

Confirms your instinct: Marsh Creek crosses from Precambrian metasedimentary rock into Idaho Batholith granite about 3.5 km below `MRS-010` (~4.5 km upstream of Big Hole) — almost exactly your proposed "halfway" point, just refined to the actual contact. Recommend **`MRS-005`** at ~44.427, -115.194.

### MFSR headwaters (below Big Hole)

```{r}
#| label: fig-map-mfsrheadwaters
map_drainage(drainage_defs[[18]])
```

```{r}
#| label: fig-mfsrheadwaters
#| results: asis
report_drainage(drainage_defs[[18]])
```

Almost entirely uniform Idaho Batholith granite (the only shift is Quaternary fill in the last ~1 km, next to `MFSR-070` — too close to be a useful new site). No lithologic case for a particular spot, so place **`MFSR-080`** near Big Hole (~44.454, -115.230, ~1.5 km below the confluence) to capture the freshly-mixed Marsh + Bear Valley signature right where the streams combine.

## Sulphur Creek

```{r}
#| label: fig-map-sulphur
map_drainage(drainage_defs[[1]])
```

```{r}
#| label: fig-sulphur
#| results: asis
report_drainage(drainage_defs[[1]])
```

Two sites: a Quaternary/Plutonic contact and a spatial anchor ~5 km below the North Fork Sulphur Creek confluence, in the upper granite section. Both moved slightly per Bryan's OnX field review — same lithologic logic, just onto an actual trail instead of an off-channel estimate.

## Pistol Creek

```{r}
#| label: fig-map-pistol
map_drainage(drainage_defs[[2]])
```

```{r}
#| label: fig-pistol
#| results: asis
report_drainage(drainage_defs[[2]])
```

Pistol Creek alternates repeatedly between granite and Quaternary valley-fill rather than crossing one clean contact — expected for a batholith-floored valley, not a crossing between different-age bedrock. The two contact-based candidates above are reasonable anchors, but expect a softer contrast than at a true bedrock-province crossing like Indian Creek (below).

The site below the Little Pistol Creek confluence has no lithologic basis (both sides Plutonic), but Little Pistol Creek supports spawning and there's no better contact to prioritize instead — a reasonable trade of geologic contrast for spawning-relevant coverage.

A second site sits right at Little Pistol Creek's own mouth — simplified from the original pick ~11 km up its own line, per Bryan's OnX review: that upstream contact wasn't reasonably reachable, and a confluence-adjacent sample still captures Little Pistol Creek's contribution for a lot less effort.

## Rapid River

```{r}
#| label: fig-map-rapid
map_drainage(drainage_defs[[3]])
```

```{r}
#| label: fig-rapid
#| results: asis
report_drainage(drainage_defs[[3]])
```

100% Idaho Batholith granite along its mapped length — geologically uniform, so the manual site is purely access-driven: a trailhead with road access near the headwaters (44.57045, -115.07117). Low priority, but easy to grab if there's time.

## Indian Creek

```{r}
#| label: fig-map-indian
map_drainage(drainage_defs[[7]])
```

```{r}
#| label: fig-indian
#| results: asis
report_drainage(drainage_defs[[7]])
```

The most geologically interesting drainage in this group: roughly half Challis volcanics, half Idaho Batholith granite. Expanded from 3 to 5 sites after Bryan revisited the spawning-location dataset and OnX field access — the lower two now bracket a narrow metasedimentary inlier directly (its downstream and upstream edges) instead of a single site sitting on it, a lower Volcanic/Plutonic contact fills in below that, and the upper site pushes further upstream for better coverage of that end of the drainage.

## Loon Creek and sub-tributaries

### Loon Creek

```{r}
#| label: fig-map-loon
p <- map_drainage(drainage_defs[[4]])
save_partner_map(p, "LOO")
p
```

```{r}
#| label: fig-loon
#| results: asis
report_drainage(drainage_defs[[4]])
```

Mostly granite (~74% Plutonic) with a real volcanic contact near its lower third and a small sedimentary patch further upstream. Expanded from 3 to 5 sites: the lower volcanic/granite contact and both edges of the sedimentary patch moved onto trail-accessible spots per Bryan's OnX review, and two new mid-drainage sites fill in the long gaps between them for better overall spatial coverage.

Note: there are two entirely separate streams named "Warm Spring Creek" in this part of the basin — one flows directly into the MFSR mainstem, the other into Loon Creek. Only the Loon Creek one (`GNIS_ID` 400163) is analyzed here.

### Warm Spring Creek

```{r}
#| label: fig-map-warmspring
p <- map_drainage(drainage_defs[[5]])
save_partner_map(p, "WSC")
p
```

```{r}
#| label: fig-warmspring
#| results: asis
report_drainage(drainage_defs[[5]])
```

### Mayfield Creek

```{r}
#| label: fig-map-mayfield
p <- map_drainage(drainage_defs[[6]])
save_partner_map(p, "MAY")
p
```

```{r}
#| label: fig-mayfield
#| results: asis
report_drainage(drainage_defs[[6]])
```

Warm Spring Creek's trunk crosses a small granite inlier surrounded by Challis volcanics. The two auto-picked candidates (13.8, 15.4 km) bracketed that same inlier only ~1.6 km apart — redundant, so they're consolidated into one site at the midpoint (14.5 km) plus a second site just above the Loon Creek confluence to anchor the drainage's downstream end.

Mayfield Creek's named trunk is short (~5 km) and 100% granite — no internal contact, so the automatic analysis skips it. But the volcanic/granite complexity on the map is real, on **East Fork** and **West Fork Mayfield Creek** (grey context lines), each crossing once between the two units — and both forks support spawning, so that variation is worth capturing.

Three manual sites: one just below the fork confluence (main-stem trunk), one on West Fork, one on East Fork. The West Fork site sits a bit upstream of the volcanic/granite contact — nudged there per Russ to sit solidly in the volcanic unit, since trunk and East Fork already cover the plutonic side; still close to a road. The East Fork site is new since Russ plans to hike up that fork anyway — moved further along the fork per Bryan's OnX review for easier access.

## Camas Creek and sub-tributaries: WF Camas & Yellowjacket creeks

### Camas Creek

```{r}
#| label: fig-map-camas
p <- map_drainage(drainage_defs[[8]])
save_partner_map(p, "CAM")
p
```

```{r}
#| label: fig-camas
#| results: asis
report_drainage(drainage_defs[[8]])
```

Expanded from 3 to 4 sites: the spatial-coverage site below the Yellowjacket confluence and the lower contact both moved onto trail-accessible spots per Bryan's OnX review, and a new mid-drainage site fills the gap before the upper contact. Everything below the Yellowjacket confluence (~6.4 km) is uniform Plutonic. Russ may hike the whole drainage to the river in September — worth grabbing the lower-reach sample while passing through rather than leaving that stretch unsampled.

### West Fork Camas Creek

```{r}
#| label: fig-map-wfcamas
map_drainage(drainage_defs[[9]])
```

```{r}
#| label: fig-wfcamas
#| results: asis
report_drainage(drainage_defs[[9]])
```

### Yellowjacket Creek

```{r}
#| label: fig-map-yellowjacket
map_drainage(drainage_defs[[10]])
```

```{r}
#| label: fig-yellowjacket
#| results: asis
report_drainage(drainage_defs[[10]])
```

100% Challis volcanic — uniform like Rapid River, so no site is recommended on lithologic grounds; place one by access if you want a point out there. (South Fork Camas Creek was dropped from this plan entirely — no salmon spawning there.)

Yellowjacket Creek stands out: the only drainage this year with a substantial share (\~46%) of Precambrian Belt Supergroup metamorphic rock alongside granite. Per `docs/review_biogeochem_tracers.qmd`, ancient metamorphic terrain is a highly radiogenic "anchor" — likely one of the most distinctive signatures in the project, worth prioritizing if time runs short.

## Wilson Creek

```{r}
#| label: fig-map-wilson
map_drainage(drainage_defs[[11]])
```

```{r}
#| label: fig-wilson
#| results: asis
report_drainage(drainage_defs[[11]])
```

Short but geologically mixed (granite, metamorphic, valley-fill) — 2 well-placed sites should capture most of the real variation. Lowest priority on Bryan's list given how few spawners use it, but worth a stop given how much lithologic variation is packed into such a short reach.

## Marble Creek

```{r}
#| label: fig-map-marble
p <- map_drainage(drainage_defs[[16]])
save_partner_map(p, "MRB")
p
```

```{r}
#| label: fig-marble
#| results: asis
report_drainage(drainage_defs[[16]])
```

The trailhead site (44.95702, -115.09996) is worth having regardless of collector — reachable by Payette NF if they're up there for Monumental, or by Bryan otherwise. Expanded from 3 to 4 sites: the two lower volcanic/granite contacts are now three, closely spaced sites near `MRB-010` — the two original contacts plus one added in between for spatial coverage — giving denser characterization of that lower stretch alongside the upper trailhead site.

## Big Creek and sub-tributaries

Note: "Rush Creek" and "Beaver Creek" are both common stream names in this basin -- the Big Creek versions used below are `GNIS_ID` 389731 and 377253 respectively (see the tributary identification reference near the top for the other streams sharing these names).

### Big Creek

```{r}
#| label: fig-map-big
p <- map_drainage(drainage_defs[[12]])
save_partner_map(p, "BIG")
p
```

```{r}
#| label: fig-big
#| results: asis
report_drainage(drainage_defs[[12]])
```

Big Creek is the most geologically diverse tributary in the plan (five rock categories, including a real metamorphic/sedimentary component) — good spatial coverage along the mainstem matters here, which is what Baxter's crew is doing.

Our two contact-based candidates (~19.4, ~45.9 km, moved slightly onto trail per Bryan's OnX review) independently land close to Hamann & Kennedy's (2012) own Big Creek samples at **site D** and **site C** — a solid cross-check on the whole approach, so D/C aren't added as separate sites. Their **sites A, B, and E** (blue triangles on the map) fill gaps our analysis missed and are added directly; a sixth site at ~30.9 km covers the largest remaining gap, in the volcanic stretch between the two contacts. That's 6 candidate sites plus the existing `BIG-010` confluence site.

### Rush Creek

```{r}
#| label: fig-map-rush
p <- map_drainage(drainage_defs[[13]])
save_partner_map(p, "RSH")
p
```

```{r}
#| label: fig-rush
#| results: asis
report_drainage(drainage_defs[[13]])
```

Remote and hard to reach beyond its lower end. Hamann & Kennedy's "Rush Cr." site (45.08826, -114.88196) sits right at the Big Creek confluence, reachable via the ranch/trail network — **priority stop**. The two contact-based candidates further upstream (6.2, 14.2 km) are geologically meaningful but secondary, unlikely to be reachable most seasons.

### Beaver Creek (Big Ck drainage)

```{r}
#| label: fig-map-beaverbig
p <- map_drainage(drainage_defs[[14]])
save_partner_map(p, "BVB")
p
```

```{r}
#| label: fig-beaverbig
#| results: asis
report_drainage(drainage_defs[[14]])
```

Same situation as Rush Creek: Hamann & Kennedy's "Beaver Cr." site (45.17340, -115.24476) sits at the Big Creek confluence — the only candidate on this Beaver Creek (`BVB`, not Marsh Creek's `BVR` — see `data/site_id_convention.md`). The two contact-based candidates further upstream (7.9, 13.7 km) are dropped — no spawning up there, so no reason to send Baxter's crew past the near-confluence stop.

### Monumental Creek

```{r}
#| label: fig-map-monumental
p <- map_drainage(drainage_defs[[15]])
save_partner_map(p, "MON")
p
```

```{r}
#| label: fig-monumental
#| results: asis
report_drainage(drainage_defs[[15]])
```

Four sites, three tiers of priority. **Lower** (two sites, both doable by Baxter's crew in one stop): right at the confluence (0 km, 45.16149, -115.1303) — Hamann & Kennedy's (2012) own "Big Cr. upstream of Monumental Cr." sample point, ~1.5 km up Big Creek from Monumental's actual mouth, Plutonic; and a second site ~1.9 miles (3.1 km) upstream on Monumental Creek itself (45.13964, -115.15895), confirmed Metasedimentary rather than sitting on the contact. Together these replace the previously-listed Hamann & Kennedy "site F" (~8 km upstream), which turns out to sit in Metasedimentary, not Plutonic. **Upper** (24.5 km, near the Monumental/Marble trailhead, 44.98076, -115.16526): Payette NF. **Middle** (13.5 km, the Volcanic/Metasedimentary contact, moved onto trail per Bryan's OnX review): geologically the most interesting site on this creek, but the hardest to reach — Payette NF if their trip extends that far, otherwise skipped this season.

## Punch list

```{r}
#| label: tbl-punchlist
punchlist <- map_dfr(drainage_defs, function(def) {
  r <- results[[def$drainage]]
  cand <- get_candidates(def)
  tibble(
    drainage = def$drainage,
    code = def$code,
    mapped_km = if (!is.null(r)) r$total_len_km else NA,
    dominant_unit = if (!is.null(r)) r$composition$lith_group[1] else NA,
    n_contacts = if (!is.null(r)) nrow(r$crossings) else NA,
    n_candidate_sites = nrow(cand),
    partner = if (!is.null(def$partner)) def$partner else "Bryan (own itinerary)"
  )
})
kable(punchlist)
```

Priorities if the season runs short on time: **Yellowjacket Creek** (only metamorphic-heavy drainage in Bryan's own list) and **Indian Creek** (cleanest single volcanic/granite contact) give the most new isotopic information per site visited. **Rapid River** is the lowest-priority stop for a *new* upstream site specifically, since it's geologically uniform — the existing `RAP-010` already characterizes it well.

## All candidate sites at a glance

Same view as the overview map above, with every candidate new site from every section overlaid -- a single quick-reference for the whole season rather than paging through each drainage individually.

```{r}
#| label: fig-map-all-candidates
#| fig-width: 9
#| fig-height: 9
#| message: false

all_candidates <- map_dfr(drainage_defs, function(def) {
  cand <- get_candidates(def)
  if (nrow(cand) == 0) return(NULL)
  cand |> filter(!is.na(lat)) |> mutate(drainage = def$drainage, code = def$code)
})
all_candidates_sf <- all_candidates |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(st_crs(fl))

p_all <- ggplot() +
  geom_sf(data = lith_clip, aes(fill = lith_group), color = NA, alpha = 0.6) +
  geom_sf(data = hu8, fill = NA, color = "grey30", linewidth = 0.5, linetype = "dashed") +
  geom_sf(data = fl, color = "grey20", linewidth = 0.15) +
  geom_sf(data = fl |> filter(GNIS_NAME == "Middle Fork Salmon River"), color = "blue", linewidth = 1) +
  geom_sf(data = target_segs, color = "blue", linewidth = 0.5) +
  geom_sf(data = existing_sites_sf, color = "black", size = 1.5) +
  geom_sf(data = all_candidates_sf, color = "red", size = 2, shape = 4, stroke = 1.3) +
  scale_fill_manual(values = litho_colors, name = "Bedrock unit") +
  map_furniture() +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right", legend.text = element_text(size = 7),
    plot.caption = element_text(hjust = 0, size = 7)
  ) +
  guides(fill = guide_legend(nrow = 5)) +
  labs(
    title = "MFSR basin: all candidate new sites, 2026 season",
    caption = str_wrap("Black dots = existing sites. Red x's = candidate new sites (this doc). Blue lines = target tributaries + named forks.", width = 100)
  )
save_partner_map(p_all, "ALL_CANDIDATES")
p_all
```

## 2026 site locations: table and KML

Every existing established site plus every candidate new site from this doc, in one place -- a single reference table for planning, and a KML file below that loads directly into OnX, Google Earth, or similar field mapping software.

```{r}
#| label: tbl-all-sites-2026
#| message: false

existing_table <- sites |>
  transmute(
    Type = "Existing",
    Drainage = drainage_group,
    `Site ID` = site_id,
    lat = lat_dd, lon = long_dd,
    Notes = paste0(waterbody_name, " -- established ", date_established)
  )

candidate_table <- all_candidates |>
  arrange(drainage, dist_upstream_km) |>
  mutate(seq = row_number(), .by = code) |>
  transmute(
    Type = "Candidate (2026)",
    Drainage = drainage,
    `Site ID` = paste0(code, "-NEW", seq),
    lat, lon,
    Notes = rationale
  )

all_sites_table <- bind_rows(existing_table, candidate_table) |> arrange(Drainage, Type)

all_sites_table |>
  select(Type, Drainage, `Site ID`, lat, lon, Notes) |>
  kable(digits = 5)
```

```{r}
#| label: export-kml
#| message: false
#| results: hide

kml_sites <- all_sites_table |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  transmute(Name = `Site ID`, Description = paste0(Drainage, " (", Type, "): ", Notes))

kml_path <- here("data/spatial/mfsr_2026_sites.kml")
if (file.exists(kml_path)) file.remove(kml_path)
st_write(kml_sites, kml_path, driver = "KML", quiet = TRUE)
```

Download: [`data/spatial/mfsr_2026_sites.kml`](../data/spatial/mfsr_2026_sites.kml) -- regenerated automatically every time this doc renders, so it always matches the table and maps above. Existing sites use their real `site_id`; candidate sites use a placeholder ID (`[CODE]-NEW1`, `[CODE]-NEW2`, ...) following the same convention as `field/partner_sample_collection_guide.qmd` -- replace with the real assigned ID once a site is actually sampled.
