---
title: "2026 Tributary Sampling Plan"
subtitle: "Geology- and GIS-informed site selection for this 2026 MFSR spawning tributaries"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(sf)
library(here)
library(knitr)
library(ggspatial)

# scale bar + north arrow, added to every map on this doc
map_furniture <- function() {
  list(
    annotation_scale(location = "bl", width_hint = 0.25, text_cex = 0.7),
    annotation_north_arrow(location = "tr", which_north = "true", height = unit(0.9, "cm"), width = unit(0.9, "cm"))
  )
}

## Colors match John's manuscript rock-type scheme (5 categories). Our raw
## USGS SGMC `GENERALIZE` field has 9 categories plus "Water" (river-channel
## digitizing artifact -- no lakes in the MFSR, so it's dropped entirely).
## Everything below groups to these 5 for both maps and tables.
litho_colors <- c(
  "Plutonic"            = "hotpink",
  "Volcanic"             = "saddlebrown",
  "Metasedimentary"      = "darkolivegreen4",
  "Sedimentary"          = "steelblue1",
  "Quaternary sediments" = "yellow"
)

lith_group_map <- c(
  "Igneous, intrusive"                             = "Plutonic",
  "Igneous, volcanic"                              = "Volcanic",
  "Igneous and Metamorphic, undifferentiated"      = "Metasedimentary",
  "Metamorphic, undifferentiated"                  = "Metasedimentary",
  "Metamorphic, sedimentary clastic"               = "Metasedimentary",
  "Metamorphic and Sedimentary, undifferentiated"  = "Metasedimentary",
  "Sedimentary, clastic"                           = "Sedimentary",
  "Sedimentary, undifferentiated"                  = "Sedimentary",
  "Unconsolidated, undifferentiated"               = "Quaternary sediments"
)

fl <- st_read(here("data/spatial/flowline_pn17_nsi_mfsr.gpkg"), quiet = TRUE)
lith <- st_read(here("data/spatial/geology/idaho_lithology.shp"), quiet = TRUE) |>
  st_make_valid() |>
  st_transform(st_crs(fl)) |>
  filter(GENERALIZE != "Water") |>
  mutate(lith_group = recode(GENERALIZE, !!!lith_group_map))
hu8 <- st_read(here("data/spatial/hu8_mfsr_dissolved.gpkg"), quiet = TRUE) |> st_transform(st_crs(fl))
sites <- read_csv(here("data/master_site_lookup.csv"), show_col_types = FALSE)
hamann <- st_read(here("data/spatial/IsoWorld.gdb"), layer = "Hamaan2012_SR_WaterSamples", quiet = TRUE) |>
  st_transform(4326)

# roads/trails for field-logistics reference on the zoomed maps
roads <- bind_rows(
  st_read(here("data/spatial/MFS_RoadsTrails_Albers96/Idaho_Roads_Clip5000_Albers96.shp"), quiet = TRUE) |> st_transform(st_crs(fl)) |> select(geometry),
  st_read(here("data/spatial/MFS_RoadsTrails_Albers96/USFS_Roads_Clip5000_Albers96.shp"), quiet = TRUE) |> st_transform(st_crs(fl)) |> select(geometry)
)
trails <- st_read(here("data/spatial/MFS_RoadsTrails_Albers96/USFS_Trails_Clip5000_Albers96.shp"), quiet = TRUE) |> st_transform(st_crs(fl))
```

Colors follow the same rock-family scheme John uses in his manuscript figures (Plutonic/Volcanic/Metasedimentary/Sedimentary/Quaternary sediments). 

This doc answers one question per target tributary: **given what we know about MFSR bedrock geology, where along this stream should a new site go to best capture real Sr variation?**

**Method**: each named tributary's NHD flowline reaches (`data/spatial/flowline_pn17_nsi_mfsr.gpkg`) are ordered upstream-to-downstream using drainage area (`TotDASqKM`, a standard proxy for position along a single named stream absent full network topology) and matched to the nearest bedrock polygon in the USGS State Geologic Map Compilation (`data/spatial/geology/idaho_lithology.shp`). This gives, for every tributary: (a) what fraction of its length drains each rock unit, and (b) where along its length it crosses from one unit into another — a **lithologic contact**, and the single best predictor of where Sr will shift most sharply. Candidate new sites are placed near contacts closest to the 1/3 and 2/3 marks of each tributary's length (so 2-3 new sites end up reasonably spread out — not evenly spaced by default, since an evenly spaced site could easily land in the middle of one uniform unit and miss the contrast entirely.

::: {.callout-note collapse="true"}
## Tributary identification reference (GNIS_ID)

Several stream names in this basin are shared by more than one physical channel, so every tributary below is selected by verified `GNIS_ID` rather than name string. `GNIS_NAME` is kept alongside for readability.

| Drainage | `GNIS_ID` used | Other stream with the same name (not used here) |
|---|---|---|
| Sulphur Creek | 391601 | 391603, an unrelated stream elsewhere in the basin |
| Warm Spring Creek | 400163 | 400162, a different "Warm Spring Creek" flowing directly into the MFSR |
| Indian Creek | 383652 | 383645, a small unrelated stream that flows into Loon Creek |
| Rush Creek | 389731 | 400063, an unrelated stream near Loon Creek |
| Beaver Creek (Big Ck) | 377253 | 377241, Marsh Creek's Beaver Creek (`BVR`) |
| Loon Creek | 399100 | -- extends ~23 km beyond the named reaches down to the true confluence |
:::

**Caveat up front**: this identifies the geologically *ideal* spot. It says nothing about whether that exact point is reachable on foot, whether it's a maintained trail, or what the current stream/trail conditions are. Treat every coordinate below as a target to aim near, not a GPS waypoint to hit exactly; adjust to the nearest safe, accessible point.

**Sample upstream or downstream of a contact?** Downstream, as a default. Water carries a rock unit's dissolved Sr signature only after it's had a chance to weather that rock, so a sample taken just *upstream* of a contact still reflects the *previous* (upstream) unit -- it hasn't touched the new one yet. Sampling just *downstream* lets the water pick up the new unit's contribution first. In practice "just downstream" should mean a bit past the contact rather than immediately at it, since water needs a short distance to mix across the channel width; a few hundred meters to a kilometer downstream is a reasonable rule of thumb where access allows. See `docs/review_biogeochem_tracers.qmd` for the underlying Sr geochemistry.

```{r}
#| label: analysis-functions
#| code-summary: "Show the tributary analysis functions"

# define drainages by verified GNIS_ID (see QA/QC callout above for how these
# were resolved). The FIRST gnis_id listed is the "trunk" used for lithology
# composition/contact-crossing/candidate-site analysis; any additional ids
# are named side-forks (e.g. Little Pistol Creek, East/West Fork Mayfield)
# that join the trunk as separate parallel channels -- shown on the map for
# context only, never merged into the trunk's up-to-down sequence (see the
# QA/QC callout on forks below for why that merge silently breaks). Where a
# tributary here is itself a named sub-tributary of ANOTHER stream in this
# list, parent_gnis_id/parent_name draws that parent stream on the zoomed map
# too, per Bryan's request to see e.g. Mayfield/Warm Spring alongside Loon.
drainage_defs <- list(
  list(
    drainage = "Sulphur Creek", 
    gnis_ids = c("391601", "387799"), 
    code = "SUL"),
  list(
    drainage = "Pistol Creek", 
    gnis_ids = c("388566", "384918"), 
    code = "PIS"),
  list(
    drainage = "Rapid River", 
    gnis_ids = c("389107"), 
    code = "RAP"),
  list(
    drainage = "Loon Creek", 
    gnis_ids = c("399100"),
    code = "LOO", 
    fill_gap = TRUE),
  list(
    drainage = "Warm Spring Creek", 
    gnis_ids = c("400163", "394337"),
    code = "WSC", 
    parent_gnis_id = "399100", 
    parent_name = "Loon Creek"),
  list(
    drainage = "Mayfield Creek", 
    gnis_ids = c("386556", "381130", "393142"), 
    code = "MAY", 
    parent_gnis_id = "399100", 
    parent_name = "Loon Creek"),
  list(
    drainage = "Indian Creek",
    gnis_ids = c("383652"), 
    code = "IND"),
  list(
    drainage = "Camas Creek",
    gnis_ids = c("378735"), 
    code = "CAM"),
  list(
    drainage = "West Fork Camas Creek",
    gnis_ids = c("393066"), 
    code = "WFC",
    parent_gnis_id = "378735",
    parent_name = "Camas Creek"),
  list(
    drainage = "Yellowjacket Creek",
    gnis_ids = c("393757", "393228"), 
    code = "YJC", 
    parent_gnis_id = "378735",
    parent_name = "Camas Creek"),
  list(
    drainage = "Wilson Creek",
    gnis_ids = c("393547"), 
    code = "WIL"),
  list(
    drainage = "Big Creek", 
    gnis_ids = c("377483"), 
    code = "BIG",
    partner = "Colden Baxter (U Idaho) and students"),
  list(
    drainage = "Rush Creek",
    gnis_ids = c("389731"), 
    code = "RSH", 
    parent_gnis_id = "377483",
    parent_name = "Big Creek"),
  list(
    drainage = "Beaver Creek (Big Ck drainage)", 
    gnis_ids = c("377253"), 
    code = "BVB", 
    partner = "Payette NF (possible)",
    parent_gnis_id = "377483", 
    parent_name = "Big Creek"),
  list(
    drainage = "Monumental Creek",
    gnis_ids = c("387122", "387739", "393151"), 
    code = "MON", 
    partner = "Payette NF (upper reach)", 
    parent_gnis_id = "377483", 
    parent_name = "Big Creek"),
  list(
    drainage = "Marble Creek", 
    gnis_ids = c("386455"), 
    code = "MRB", 
    partner = "Payette NF (unconfirmed -- Bryan likely collects if they don't go in)", 
    parent_gnis_id = "377483",
    parent_name = "Big Creek"),
  # headwaters reach: Big Hole (Marsh Ck + Bear Valley Ck confluence, the
  # official start of the MFSR) down to the Boundary Creek put-in -- Bryan
  # already has upper Marsh/Bear Valley samples but nothing in these two
  # currently-unsampled stretches. Restricted to a TotDASqKM window on an
  # already-resolved single-GNIS_ID stream, so no gap-fill needed.
  list(
    drainage = "Lower Marsh Creek (MRS-010 to Big Hole)", 
    gnis_ids = c("386480"), 
    code = "MRS", 
    totda_range = c(360, 386)),
  list(
    drainage = "MFSR headwaters (Big Hole to Boundary Creek)", 
    gnis_ids = c("386841"), 
    code = "MFSR", 
    totda_range = c(884, 1022))
)

# directed downstream walk to absorb unnamed "gap" reaches (e.g. Loon Creek's
# missing lower third) without wandering onto side tributaries: at each step,
# only the touching reach with the LARGEST TotDASqKM is the true trunk
# continuation, and we stop as soon as that reach turns out to be a real,
# differently-named waterbody (the actual confluence). IMPORTANT: only
# Loon Creek actually has a confirmed attribute gap -- applying this to every
# tributary is itself a bug we hit (Warm Spring Creek's downstream walk from
# its own mouth immediately entered Loon Creek's unnamed gap and kept
# absorbing Loon Creek's own reaches, since fill_downstream_gap has no way to
# know it had already crossed into a different stream's territory once both
# are unnamed). So this is now opt-in per drainage (`fill_gap = TRUE`), set
# only where a real gap was confirmed against a known `-010` site -----
fill_downstream_gap <- function(seed_ids, fl, max_iter = 60) {
  named <- fl |> filter(GNIS_ID %in% seed_ids)
  current <- named[which.max(named$TotDASqKM), ]
  pool <- fl |> filter(COMID != current$COMID[1])
  filled_ids <- c()
  for (i in seq_len(max_iter)) {
    touching <- suppressMessages(st_filter(pool, current, .predicate = st_is_within_distance, dist = 1))
    touching <- touching |> filter(TotDASqKM > current$TotDASqKM[1])
    if (nrow(touching) == 0) break
    nxt <- touching[which.max(touching$TotDASqKM), ]
    if (!is.na(nxt$GNIS_NAME)) break # reached a real, different named waterbody -- stop
    filled_ids <- c(filled_ids, nxt$COMID)
    pool <- pool |> filter(COMID != nxt$COMID)
    current <- nxt
  }
  bind_rows(named, fl[fl$COMID %in% filled_ids, ])
}

# analysis functions ----

# fetch reaches for a set of GNIS_IDs, honoring totda_range/fill_gap the same
# way regardless of whether it's the trunk or a fork
fetch_reaches <- function(ids, def) {
  if (length(ids) == 0) return(fl[0, ])
  if (!is.null(def$totda_range)) {
    fl |> filter(GNIS_ID %in% ids, TotDASqKM >= def$totda_range[1], TotDASqKM <= def$totda_range[2])
  } else if (isTRUE(def$fill_gap)) {
    fill_downstream_gap(ids, fl)
  } else {
    fl |> filter(GNIS_ID %in% ids)
  }
}

# order by TotDASqKM, join lithology, compute cumulative distance -- using
# st_point_on_surface (guaranteed to sit ON the line) rather than st_centroid
# (which for a bent/curvy reach can land off the line entirely -- the direct
# cause of candidate markers not landing on the stream line)
add_lith_and_order <- function(x) {
  if (nrow(x) == 0) return(x)
  x <- x |> arrange(TotDASqKM)
  pts <- suppressWarnings(st_point_on_surface(x))
  x$lith_group <- st_join(pts, lith["lith_group"], join = st_nearest_feature)$lith_group
  x$cumlen <- cumsum(x$LENGTHKM)
  ll <- st_transform(pts, 4326)
  x$lat <- st_coordinates(ll)[, 2]
  x$lon <- st_coordinates(ll)[, 1]
  x
}

analyze_trib <- function(def) {
  primary_id <- def$gnis_ids[1]
  fork_ids <- setdiff(def$gnis_ids, primary_id)

  # trunk only: composition, contact-crossings, and candidate sites all come
  # from this single ordered line. Named side-forks (fork_ids) are fetched
  # separately and never merged into the trunk's TotDASqKM sequence -- doing
  # that was the bug behind off-line candidates on Pistol Creek (vs. Little
  # Pistol Creek) and Mayfield Creek (vs. its East/West Forks): forks are
  # parallel channels whose drainage areas interleave with the trunk's rather
  # than continuing it, so sorting them together produces "contacts" between
  # geometrically unrelated points on two different lines.
  seg <- fetch_reaches(primary_id, def)
  seg_forks <- fetch_reaches(fork_ids, def)
  if (nrow(seg) == 0) return(NULL)

  seg <- add_lith_and_order(seg)
  seg_forks <- add_lith_and_order(seg_forks)

  total_len <- sum(seg$LENGTHKM)
  comp <- st_drop_geometry(seg) |>
    summarise(len = sum(LENGTHKM), .by = lith_group) |>
    mutate(pct = round(100 * len / total_len, 1)) |>
    arrange(desc(pct))

  d <- st_drop_geometry(seg)
  changes <- which(d$lith_group[-1] != d$lith_group[-nrow(d)])
  crossings <- tibble()
  if (length(changes) > 0) {
    crossings <- map_dfr(changes, function(i) {
      tibble(
        from_unit = d$lith_group[i], to_unit = d$lith_group[i + 1],
        cumlen_km = round((d$cumlen[i] + d$cumlen[i + 1]) / 2, 1),
        lat = round((d$lat[i] + d$lat[i + 1]) / 2, 5),
        lon = round((d$lon[i] + d$lon[i + 1]) / 2, 5)
      )
    })
  }

  list(
    drainage = def$drainage, code = def$code, partner = def$partner,
    total_len_km = round(total_len, 1), composition = comp, crossings = crossings,
    seg = seg, seg_forks = seg_forks, fork_ids = fork_ids,
    parent_gnis_id = def$parent_gnis_id, parent_name = def$parent_name
  )
}

pick_candidates <- function(r, fracs = c(0.33, 0.66)) {
  if (is.null(r) || nrow(r$crossings) == 0) return(tibble())
  total_len <- r$total_len_km
  tol <- max(4, total_len * 0.15)
  map_dfr(fracs, function(f) {
    target <- total_len * (1 - f)
    idx <- which.min(abs(r$crossings$cumlen_km - target))
    best <- r$crossings[idx, ]
    if (abs(best$cumlen_km - target) <= tol) {
      tibble(
        dist_upstream_km = round(total_len - best$cumlen_km, 1),
        lat = best$lat, lon = best$lon,
        rationale = paste0(best$from_unit, " <-> ", best$to_unit),
        source = "Analysis"
      )
    } else {
      tibble(
        dist_upstream_km = round(total_len * f, 1), lat = NA, lon = NA,
        rationale = "no major contact nearby - place by even spacing / access",
        source = "Analysis"
      )
    }
  }) |> distinct(lat, lon, .keep_all = TRUE)
}

# unlike pick_candidates() (which selects 2-3 well-spaced NEW site spots),
# this surfaces EVERY detected contact -- used for the Bear Valley/Marsh
# Creek retrospective review below, where the question is "how do existing
# sites compare to all real contacts," not "where should new sites go"
all_crossings <- function(r) {
  if (is.null(r) || nrow(r$crossings) == 0) return(tibble())
  r$crossings |>
    transmute(
      dist_upstream_km = round(r$total_len_km - cumlen_km, 1),
      lat, lon,
      rationale = paste0(from_unit, " <-> ", to_unit),
      source = "Analysis"
    )
}

# combine the automatic contact-based picks with any hand-picked sites Bryan
# adds as the season firms up -- both kept side by side (tagged by `source`)
# rather than one replacing the other, unless noted otherwise for a specific
# drainage (e.g. Warm Spring Creek's two close-together auto picks were
# consolidated into one at Bryan's request)
manual_site <- function(dist_upstream_km, lat, lon, rationale) {
  tibble(dist_upstream_km = dist_upstream_km, lat = lat, lon = lon, rationale = rationale, source = "Manual")
}

results <- map(drainage_defs, analyze_trib)
names(results) <- map_chr(drainage_defs, "drainage")

# Bear Valley Creek and upper Marsh Creek (above MRS-010) are a retrospective
# check on 2024 site placement, not 2026 targets -- kept OUT of drainage_defs
# on purpose so they don't show up in the punch list or the "all candidate
# sites this season" map, but still get the same analyze_trib()/report_
# drainage()/map_drainage() treatment by appending directly into `results`
bvc_def <- list(drainage = "Bear Valley Creek", gnis_ids = c("377200"), code = "BVC")
mrs_upper_def <- list(drainage = "Marsh Creek (upper, above MRS-010)", gnis_ids = c("386480"), code = "MRS", totda_range = c(0, 360))
results[[bvc_def$drainage]] <- analyze_trib(bvc_def)
results[[mrs_upper_def$drainage]] <- analyze_trib(mrs_upper_def)

# manual overrides for cases where the generic 1/3-2/3-contact heuristic
# isn't the right call (see the headwaters section) -- keyed by drainage name
# so both report_drainage() and map_drainage() (and the final summary map)
# all read the same recommended site rather than the map showing one thing
# and the table showing another
manual_overrides <- list(
  "Lower Marsh Creek (MRS-010 to Big Hole)" = manual_site(
    4.5, 44.427, -115.194,
    "Metasedimentary <-> Plutonic ('halfway' point lands almost exactly on this contact)"
    ),
  "MFSR headwaters (Big Hole to Boundary Creek)" = manual_site(
    16.2, 44.454, -115.230,
    "just below Big Hole -- post-confluence mixed signature"
    ),
  "Sulphur Creek" = bind_rows(
    pick_candidates(results[["Sulphur Creek"]]),
    manual_site(
      14.6, 44.54938, -115.4029, 
      "~5 km downstream of the North Fork Sulphur Creek confluence")
  ),
  "Pistol Creek" = bind_rows(
    pick_candidates(results[["Pistol Creek"]]),
    manual_site(
      4, 44.72139, -115.1921,
      "just below the Little Pistol Creek confluence"),
    manual_site(NA, 44.72969, -115.22054, "on Little Pistol Creek itself")),
  "Rapid River" = manual_site(20.3, 44.57045, -115.07117,"trailhead with road access"),
  "Indian Creek" = bind_rows(
    pick_candidates(results[["Indian Creek"]]),
    manual_site(6.0, 44.79394, -115.1325, "narrow metasedimentary inlier")),
  "Marble Creek" = bind_rows(
    pick_candidates(results[["Marble Creek"]]),
    manual_site(34.2, 44.95702, -115.09996, "top of drainage by trailhead"),
    manual_site(7.4, 44.77893, -114.9775, "second contact downstream of 10.9 km candidate")),
  "Loon Creek" = bind_rows(
    pick_candidates(results[["Loon Creek"]]),
    manual_site(39.8, 44.54617, -114.85635, "other edge sedimentary patch")),
  # replaces (not adds to) the two auto-picked candidates, which bracketed the
  # same small granite inlier only ~1.6 km apart -- consolidated to one
  # in-between site plus one near the Loon Creek confluence, per Bryan
  "Warm Spring Creek" = bind_rows(
    manual_site(14.5, 44.61547, -114.5958, "midpoint of the granite inlier (averages the two closely-spaced auto picks)"),
    manual_site(1.5, 44.65062, -114.7171, "just upstream of the confluence with Loon Creek")
  ),
  "Mayfield Creek" = bind_rows(
    manual_site(4.6, 44.54007, -114.8015, "just below the East/West Fork confluence, on the short main-stem trunk"),
    manual_site(NA, 44.50463, -114.7439, "on West Fork Mayfield Creek, just upstream of its volcanic/granite contact -- moved a bit further upstream (from the contact itself) per Russ, to sit solidly in the volcanic unit rather than right on the boundary; the trunk and East Fork sites already cover the plutonic side"),
    manual_site(NA, 44.55133, -114.74546, "on East Fork Mayfield Creek -- Russ wants to hike up the East Fork a bit, so worth a stop while he's up there")
  ),
  "Monumental Creek" = bind_rows(
    pick_candidates(results[["Monumental Creek"]]),
    manual_site(24.5, 44.98076, -115.16526, "near the Monumental/Marble trailhead, upper drainage"),
    manual_site(7.9, 45.09937, -115.15733, "Hamann & Kennedy (2012) 'Monumental Cr. (site F)' -- lower-drainage, reachable from the Big Creek side; complements the upper trailhead site above, since the analysis-picked mid-drainage contact (14.9 km) is the least accessible of the three")
  ),
  # Hamann & Kennedy (2012) sites snap almost exactly (0 m) onto our resolved
  # trunk line -- sites D and C aren't added here since they're already
  # closely matched by the existing 20.5/45 km auto candidates (see prose
  # below); E, A, and B fill in gaps the auto analysis didn't pick
  "Big Creek" = bind_rows(
    pick_candidates(results[["Big Creek"]]),
    manual_site(8.7, 45.10261, -114.84607, "Hamann & Kennedy (2012) 'Lower Big Cr.' (site E)"),
    manual_site(60.9, 45.10630, -115.32839, "Hamann & Kennedy (2012) 'Upper Big Cr.' (site A)"),
    manual_site(58.2, 45.12779, -115.31390, "Hamann & Kennedy (2012) 'Big Cr. downstream of Logan Cr.' (site B)"),
    manual_site(30.9, 45.14177, -115.05557, "mid-drainage spatial coverage within the volcanic stretch between the 20.5 km and 45 km contact-based candidates")
  ),
  # Camas Creek's two auto-picked candidates (15, 38 km) sit well upstream of
  # the Yellowjacket confluence (~6.4 km) -- the whole stretch below that is
  # uniform Plutonic, so this manual site is for spatial coverage ahead of
  # the confluence, not a lithologic contact
  "Camas Creek" = bind_rows(
    pick_candidates(results[["Camas Creek"]]),
    manual_site(3.5, 44.89030, -114.6783, "between CAM-010 and the Yellowjacket Creek confluence (~6.4 km) -- lithologically uniform here (Plutonic), so this is for spatial coverage; per Russ, worth grabbing given how much variation Camas and Yellowjacket show overall")
  ),
  "Rush Creek" = bind_rows(
    pick_candidates(results[["Rush Creek"]]),
    manual_site(0.0, 45.08826, -114.88196, "Hamann & Kennedy (2012) 'Rush Cr.' reference site -- near the Big Creek confluence and reachable via the ranch/trail network; priority stop since everything upstream is remote with no road")
  ),
  "Beaver Creek (Big Ck drainage)" = bind_rows(
    pick_candidates(results[["Beaver Creek (Big Ck drainage)"]]),
    manual_site(0.0, 45.17340, -115.24476, "Hamann & Kennedy (2012) 'Beaver Cr.' reference site -- near the Big Creek confluence; priority stop, same access logic as Rush Creek")
  ),
  # retrospective review, not new-site picks -- show every detected contact
  # (all_crossings()) rather than the 2-3-site spacing heuristic
  "Bear Valley Creek" = all_crossings(results[["Bear Valley Creek"]]),
  "Marsh Creek (upper, above MRS-010)" = all_crossings(results[["Marsh Creek (upper, above MRS-010)"]])
)

get_candidates <- function(def) {
  if (!is.null(manual_overrides[[def$drainage]])) return(manual_overrides[[def$drainage]])
  pick_candidates(results[[def$drainage]])
}

# composition/candidate table condensed to one line + one compact table (no
# lat/lon columns -- those are only needed for the map marker, not for
# scanning the text) so each tributary section reads as map -> table -> prose
# rather than map -> table -> table -> prose
report_drainage <- function(def, candidate_label = "Candidate new site positions") {
  r <- results[[def$drainage]]
  if (is.null(r)) {
    cat("No flowline match found for", def$drainage, "-- check GNIS_ID.\n\n")
    return(invisible())
  }
  cand <- get_candidates(def)
  comp_str <- r$composition |>
    mutate(part = sprintf("%.0f%% %s", pct, lith_group)) |>
    pull(part) |>
    paste(collapse = ", ")
  cat(sprintf("**%s (`%s`)** -- %.1f km of mapped channel. Bedrock: %s.\n\n", def$drainage, def$code, r$total_len_km, comp_str))
  if (nrow(cand) > 0) {
    cat(candidate_label, " (distance upstream of the `-010` confluence site):\n\n", sep = "")
    print(kable(cand |> select(source, dist_upstream_km, rationale) |>
      rename(Source = source, `km upstream` = dist_upstream_km, Rationale = rationale)))
  } else {
    cat("No lithologic contact detected along the mapped channel -- geologically uniform; ",
        "space new sites evenly based on access rather than geology.\n", sep = "")
  }
  if (length(r$fork_ids) > 0 && nrow(r$seg_forks) > 0) {
    cat("\nNamed side-fork(s) also shown on the map below (grey lines) as context: ",
        paste(unique(r$seg_forks$GNIS_NAME), collapse = ", "), ".\n", sep = "")
  }
  cat("\n")
}

# zoomed per-tributary map: bedrock context + the (corrected, gap-filled)
# tributary line + existing site(s) + candidate new site(s). Candidates come
# from get_candidates(), which reads manual_overrides where set (e.g. the
# headwaters reach) so the map always matches what report_drainage() prints
# -----
map_drainage <- function(def) {
  r <- results[[def$drainage]]
  if (is.null(r)) return(invisible(NULL))
  seg <- r$seg
  cand <- get_candidates(def)

  # zoom extent covers the trunk AND any named forks, so the confluence with
  # a fork (e.g. Little Pistol Creek, East/West Fork Mayfield) is visible
  zoom_geom <- if (nrow(r$seg_forks) > 0) {
    st_union(st_union(seg), st_union(r$seg_forks))
  } else {
    st_union(seg)
  }
  bbox <- st_bbox(st_buffer(zoom_geom, 3000))
  bbox_poly <- st_as_sfc(bbox)
  lith_zoom <- suppressWarnings(st_crop(lith, bbox))

  # coord_sf's xlim/ylim only set the *view*; when the bbox's aspect ratio
  # doesn't match the figure device, ggplot expands one axis to preserve true
  # geographic aspect, and anything well outside the intended zoom (other
  # sites, the whole basin's stream network) can end up inside that expanded
  # view. Crop every layer to the bbox explicitly -- with nothing left
  # outside the zoom to leak in, and expand = FALSE below to stop coord_sf
  # padding the view further, the map is a true zoom regardless of aspect.
  fl_zoom <- suppressWarnings(st_crop(fl, bbox))
  existing <- sites |>
    filter(stream_code == r$code) |>
    st_as_sf(coords = c("long_dd", "lat_dd"), crs = 4326) |>
    st_transform(st_crs(fl)) |>
    st_filter(bbox_poly, .predicate = st_intersects)
  roads_zoom <- suppressWarnings(st_crop(roads, bbox))
  trails_zoom <- suppressWarnings(st_crop(trails, bbox))
  hamann_zoom <- suppressWarnings(st_crop(st_transform(hamann, st_crs(fl)), bbox))
  hu8_zoom <- suppressWarnings(st_crop(hu8, bbox))

  p <- ggplot() +
    geom_sf(data = lith_zoom, aes(fill = lith_group), color = NA, alpha = 0.75) +
    geom_sf(data = fl_zoom, color = "grey50", linewidth = 0.15)

  if (nrow(hu8_zoom) > 0) p <- p + geom_sf(data = hu8_zoom, fill = NA, color = "grey30", linewidth = 0.6, linetype = "dashed")

  if (nrow(roads_zoom) > 0) p <- p +
    geom_sf(data = roads_zoom, color = "white", linewidth = 0.7) +
    geom_sf(data = roads_zoom, color = "firebrick", linewidth = 0.35)
  if (nrow(trails_zoom) > 0) p <- p +
    geom_sf(data = trails_zoom, color = "white", linewidth = 0.6) +
    geom_sf(data = trails_zoom, color = "black", linewidth = 0.3, linetype = "22")
  if (nrow(hamann_zoom) > 0) p <- p + geom_sf(data = hamann_zoom, color = "blue", size = 2, shape = 2, stroke = 1)

  # parent stream (e.g. Loon Creek behind Warm Spring/Mayfield), cropped to
  # this same zoom -- typically shows just the confluence-adjacent stretch
  title_suffix <- ""
  if (!is.null(r$parent_gnis_id)) {
    parent_zoom <- suppressWarnings(st_crop(fl |> filter(GNIS_ID == r$parent_gnis_id), bbox))
    if (nrow(parent_zoom) > 0) {
      p <- p + geom_sf(data = parent_zoom, color = "black", linewidth = 1, linetype = "dotted", alpha = 0.7)
      title_suffix <- paste0(" -- tributary of ", r$parent_name)
    }
  }
  # named side-forks (e.g. Little Pistol Creek), shown but not analyzed
  if (nrow(r$seg_forks) > 0) {
    p <- p + geom_sf(data = r$seg_forks, color = "grey35", linewidth = 0.6)
  }

  p <- p +
    geom_sf(data = seg, color = "steelblue4", linewidth = 1.1) +
    scale_fill_manual(values = litho_colors, name = "Bedrock unit", drop = TRUE) +
    map_furniture() +
    coord_sf(xlim = bbox[c("xmin", "xmax")], ylim = bbox[c("ymin", "ymax")], expand = FALSE) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position = "right", legend.text = element_text(size = 7), legend.title = element_text(size = 8),
      plot.caption = element_text(hjust = 0, size = 7)
    ) +
    labs(title = paste0(def$drainage, " (`", def$code, "`)", title_suffix), x = NULL, y = NULL)

  if (nrow(existing) > 0) {
    p <- p +
      geom_sf(data = existing, color = "black", size = 2.5, shape = 17) +
      geom_sf_text(data = existing, aes(label = site_id), size = 2.6, fontface = "bold", nudge_y = 400, check_overlap = TRUE)
  }
  if (nrow(cand) > 0 && any(!is.na(cand$lat))) {
    cand_sf <- cand |>
      filter(!is.na(lat)) |>
      st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
      st_transform(st_crs(fl))
    p <- p + geom_sf(data = cand_sf, color = "red", size = 3, shape = 4, stroke = 1.5)
  }
  p +
    labs(caption = str_wrap(paste0(
      "Black triangle = existing site (labeled). Red x = candidate new site. Grey dashed outline = HUC8 basin edge. ",
      "Red line (white-cased) = road. Black dashed line (white-cased) = trail. Blue triangle = Hamann & Kennedy (2012) reference site.",
      if (nrow(r$seg_forks) > 0) " Grey line = named side-fork (context only)." else "",
      if (title_suffix != "") " Black dotted stream line = parent stream." else ""
    ), width = 100))
}
```

## Overview map

```{r}
#| label: fig-overview-map
#| fig-width: 9
#| fig-height: 9
#| message: false

lith_clip <- st_intersection(lith, st_union(hu8))
existing_sites_sf <- sites |>
  st_as_sf(coords = c("long_dd", "lat_dd"), crs = 4326) |>
  st_transform(st_crs(fl))

# uses the corrected, GNIS_ID-verified + gap-filled tributary geometries
# (results[[...]]$seg), not a raw GNIS_NAME filter -- avoids re-introducing
# the name-collision/gap bugs described in the QA/QC callout above
target_segs <- bind_rows(map(results, function(x) {
  if (is.null(x)) return(NULL)
  bind_rows(x$seg, x$seg_forks)
}))

ggplot() +
  geom_sf(data = lith_clip, aes(fill = lith_group), color = NA, alpha = 0.6) +
  geom_sf(data = hu8, fill = NA, color = "grey30", linewidth = 0.5, linetype = "dashed") +
  geom_sf(data = fl, color = "grey20", linewidth = 0.15) +
  geom_sf(data = fl |> filter(GNIS_NAME == "Middle Fork Salmon River"), color = "blue", linewidth = 1) +
  geom_sf(data = target_segs, color = "blue", linewidth = 0.5) +
  geom_sf(data = hamann, color = "red", size = 1, shape = 4) +
  geom_sf(data = existing_sites_sf, color = "black", size = 1.5) +
  scale_fill_manual(values = litho_colors, name = "Bedrock unit") +
  map_furniture() +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right", legend.text = element_text(size = 7),
    plot.caption = element_text(hjust = 0, size = 7)
  ) +
  guides(fill = guide_legend(nrow = 5)) +
  labs(
    title = "MFSR basin: bedrock lithology and sample sites",
    caption = str_wrap("Black dots = existing sample sites (master_site_lookup.csv). Red x's = Hamann & Kennedy (2012) Big Creek water samples.", width = 100)
  )
```

Bedrock is dominated by two igneous provinces of very different age (Cretaceous Idaho Batholith granite and Eocene Challis Volcanics/intrusives), with a real Precambrian metamorphic component in the northeast part of the basin (Yellowjacket, Big Creek headwaters area) — see `docs/review_biogeochem_tracers.qmd` for the full area-weighted breakdown and why this predicts sharp, mappable Sr contrasts in exactly those tributaries that cross from one province into another.

## Who's covering what — 2026 season plan

Every drainage in this plan is analyzed the same way regardless of who ends up collecting it (the geology doesn't stop mattering just because someone else is out there) — but it's worth laying out who's doing what and when, both to keep partner coordination straight and to sequence Bryan's own remaining trips sensibly.

**Colden Baxter (University of Idaho) and students** will collect Big Creek this summer, plus the near-confluence sites on Beaver Creek, Monumental Creek, and Rush Creek where those sub-tributaries meet Big Creek (see each drainage's section below for exact coordinates). Rush Creek beyond that near-confluence spot is remote and difficult, so realistically just the one sample there. The candidate site coordinates below are exactly what should go into `field/partner_sample_collection_guide.qmd` for their crew.

**Payette National Forest biologists** will help with upper Monumental Creek, and may also be able to cover upper Beaver Creek (otherwise Bryan). Upper Marble Creek is hopefully also Payette — if that pans out, Bryan may fly into lower Marble Creek to pick up the two lower-drainage sites there instead of hiking the whole thing.

**Russ** will collect Loon Creek, Warm Spring Creek, and Mayfield Creek next week — see the new manual sites added to those three drainages below, all requested directly by Russ ahead of that trip. He'll likely do Camas Creek in September, hiking down the drainage to the river and floating out, which is also why a manual site was added between `CAM-010` and the Yellowjacket confluence — worth grabbing given how much Sr variation both Camas and Yellowjacket show.

**Bryan's own remaining itinerary**: Sulphur Creek, Rapid River, (possibly additional sites on Marsh Creek/Bear Valley Creek beyond the 2024 sites and this doc's headwaters review), Pistol Creek, Indian Creek, lower Marble Creek, and Yellowjacket Creek. Rough trip logistics:

- **Rapid River** is an easy drive-in, grab-and-leave stop, and could be combined with the Sulphur Creek trip (and possibly extra Marsh/Bear Valley sites) in one outing.
- **Sulphur Creek** is road-accessible but far enough to be a long day trip or a short overnight based out of Boundary Creek campground.
- **Pistol Creek and Indian Creek** call for a 4-5 day trip — in through Pistol, out through Indian, with a shuttle needed.
- **Lower Marble Creek** is a likely fly-in, especially if Payette covers upper Marble.
- **Yellowjacket Creek** should be doable via road access.

## Headwaters

The Middle Fork Salmon River officially begins at "Big Hole," where Marsh Creek and Bear Valley Creek meet (confirmed in the flowline data: both streams' mouths sit exactly at the same point, 0 m apart, which is also the most-upstream end of the "Middle Fork Salmon River" `GNIS_NAME` reaches). Bryan already has samples from upper Bear Valley Creek (`BVC-010/020/030`) and upper Marsh Creek (`MRS-010/020/030`) -- the first two subsections below retroactively run the same contact-detection analysis used everywhere else in this doc against those *already-established* 2024 sites, as a check on whether they landed somewhere geologically useful. Beyond that, two stretches remain genuinely unsampled: the lower ~8 km of Marsh Creek between `MRS-010` and Big Hole itself, and the entire ~18 km of brand-new MFSR channel from Big Hole down to the existing `MFSR-070` site at the Boundary Creek put-in.

### Bear Valley Creek (existing sites review)

```{r}
#| label: fig-map-bvc
map_drainage(bvc_def)
```

```{r}
#| label: fig-bvc
#| results: asis
report_drainage(bvc_def, candidate_label = "Detected lithologic contacts (for comparison against existing sites)")
```

Bear Valley Creek is mostly Quaternary valley-fill (the broad glaciated meadow the creek is named for) with a real Plutonic component, and the analysis finds four contacts along its length. Comparing the four red x's to the three black triangles: `BVC-010` and `BVC-030` both sit within a couple km of a real contact -- good placement, not just even spacing. `BVC-020`, in the middle, sits well inside a uniform Quaternary stretch with no contact nearby, so it's likely doing more for spatial/access coverage than for capturing a lithologic transition -- not a bad site, just not one landing on a contrast. More notable: the westernmost (most-upstream) contact sits beyond `BVC-030` entirely, past the current sampling extent -- if Bear Valley Creek gets revisited, that's the one gap in the existing 2024 placement worth a look.

### Marsh Creek, above MRS-010 (existing sites review)

```{r}
#| label: fig-map-mrsupper
map_drainage(mrs_upper_def)
```

```{r}
#| label: fig-mrsupper
#| results: asis
report_drainage(mrs_upper_def, candidate_label = "Detected lithologic contacts (for comparison against existing sites)")
```

This reach comes back 100% Quaternary sediments -- no internal contact at all above `MRS-010`. That's a genuinely useful negative result: there's no lithologic argument that `MRS-010/020/030` missed anything, because there's nothing to miss until the Metasedimentary/Plutonic contact identified in Lower Marsh Creek below (which sits *downstream* of `MRS-010`, not in this reach). The tight 0.708 cluster these sites fall into (see `eda_water_sr.qmd`) is consistent with sampling a uniform valley-fill unit rather than a sign of undersampling.

### Lower Marsh Creek

```{r}
#| label: fig-map-lowermarsh
map_drainage(drainage_defs[[17]])
```

```{r}
#| label: fig-lowermarsh
#| results: asis
report_drainage(drainage_defs[[17]])
```

This matches what you'd expect from the review doc's geology section, and it's a clean confirmation of your own instinct: Marsh Creek runs through Precambrian metasedimentary rock below Beaver Creek/Cape Horn Creek, then crosses into Idaho Batholith granite about 3.5 km downstream of `MRS-010` (roughly 4.5 km upstream of Big Hole) -- almost exactly the "halfway" point you proposed, just refined to the actual contact. Recommend **`MRS-005`** at approximately 44.427, -115.194.

### MFSR headwaters (below Big Hole)

```{r}
#| label: fig-map-mfsrheadwaters
map_drainage(drainage_defs[[18]])
```

```{r}
#| label: fig-mfsrheadwaters
#| results: asis
report_drainage(drainage_defs[[18]])
```

Unlike lower Marsh Creek, this reach is almost entirely uniform Idaho Batholith granite start to finish (the only shift is to Quaternary valley-fill in the last ~1 km, right next to `MFSR-070` -- not a useful *new* site since it's basically on top of the existing one). There's no lithologic argument for placing this new site at a particular spot along the 18 km, so the recommendation is driven by the sampling question itself rather than geology: place **`MFSR-080`** close to Big Hole (roughly 44.454, -115.230, ~1.5 km below the confluence) to capture the newly-mixed Marsh Creek + Bear Valley Creek signature right where the two streams combine, rather than partway down a reach that won't change much either way.

## Sulphur Creek

```{r}
#| label: fig-map-sulphur
map_drainage(drainage_defs[[1]])
```

```{r}
#| label: fig-sulphur
#| results: asis
report_drainage(drainage_defs[[1]])
```

The manual site (~5 km downstream of the North Fork Sulphur Creek confluence) sits between that confluence and the granite/Quaternary contact identified by the analysis, so it adds a spatial anchor in the upper granite section without duplicating the contact site.

## Pistol Creek

```{r}
#| label: fig-map-pistol
map_drainage(drainage_defs[[2]])
```

```{r}
#| label: fig-pistol
#| results: asis
report_drainage(drainage_defs[[2]])
```

Pistol Creek's channel alternates repeatedly between granite and valley-fill (Quaternary alluvium) rather than crossing a single clean contact — expected given it's flowing through a broad batholith-floored valley, not between two different bedrock provinces. The two candidate positions above are still reasonable anchor points, but don't expect as sharp a contrast here as in a drainage that crosses between actual different-age bedrock (e.g. Indian Creek, below).

On the manual site just below the Little Pistol Creek confluence: this makes sense to add. There's no lithologic reason for it (both sides are Plutonic), but Little Pistol Creek supports spawning and there's no clean contact anywhere else on Pistol Creek to prioritize instead, so a site here trades geologic contrast for biologically relevant spatial coverage right where a significant spawning tributary joins -- a reasonable trade given the alternative is no extra site at all.

A second site now goes on Little Pistol Creek itself, ~11 km up its own line (not the Pistol Creek trunk), right at a real Quaternary sediments/Plutonic contact and close to a trail -- unlike the confluence site above, this one does have a lithologic reason, and per Russ it's worth the stop while he's up that far anyway.

## Rapid River

```{r}
#| label: fig-map-rapid
map_drainage(drainage_defs[[3]])
```

```{r}
#| label: fig-rapid
#| results: asis
report_drainage(drainage_defs[[3]])
```

Rapid River is 100% Idaho Batholith granite along its entire mapped length — geologically uniform. There's no lithologic reason to place a new site at any particular spot, so the manual site here is purely access-driven: a trailhead with road access near the drainage headwaters (44.57045, -115.07117). Low priority given the uniform geology, but easy to reach if there's time.

## Indian Creek

```{r}
#| label: fig-map-indian
map_drainage(drainage_defs[[7]])
```

```{r}
#| label: fig-indian
#| results: asis
report_drainage(drainage_defs[[7]])
```

Indian Creek is the most geologically interesting of this group: it's roughly half Challis volcanics, half Idaho Batholith granite, with the contact sitting well inside the drainage rather than at either end — a real candidate for the single sharpest Sr contrast among all the tributaries sampled this way (see also the Warm Spring/Mayfield discussion below, which shows the same volcanic/granite juxtaposition even more repeatedly). The manual site adds a third stop at a narrow metasedimentary sliver between the two existing contacts -- a small feature, but it's a different rock family entirely (not just another volcanic/granite flip), so worth the extra stop if you're already walking this stretch.

## Loon Creek and sub-tributaries

### Loon Creek

```{r}
#| label: fig-map-loon
map_drainage(drainage_defs[[4]])
```

```{r}
#| label: fig-loon
#| results: asis
report_drainage(drainage_defs[[4]])
```

Loon Creek is mostly granite (Plutonic, ~74%) with a real volcanic contact near its lower third and a small sedimentary patch further upstream. The added manual site (now 39.8 km upstream, nudged slightly further upstream from the original 39.2 km per Russ) sits at the other edge of that same sedimentary patch from the existing 36.1 km candidate, so together they bracket the whole patch rather than just clipping one edge of it.

Note: there are two entirely separate streams named "Warm Spring Creek" in this part of the basin — one flows directly into the MFSR mainstem, the other into Loon Creek. Only the Loon Creek one (`GNIS_ID` 400163) is analyzed here.

### Warm Spring Creek

```{r}
#| label: fig-map-warmspring
map_drainage(drainage_defs[[5]])
```

```{r}
#| label: fig-warmspring
#| results: asis
report_drainage(drainage_defs[[5]])
```

### Mayfield Creek

```{r}
#| label: fig-map-mayfield
map_drainage(drainage_defs[[6]])
```

```{r}
#| label: fig-mayfield
#| results: asis
report_drainage(drainage_defs[[6]])
```

Warm Spring Creek's named trunk crosses a small granite inlier surrounded by Challis volcanics. The original two auto-picked candidates (13.8 and 15.4 km upstream) both bracketed that same small inlier only ~1.6 km apart -- redundant for a feature that small, so per Bryan's call they're consolidated into one site at the midpoint (14.5 km upstream) plus a second site just upstream of the confluence with Loon Creek, to also anchor the downstream end of the drainage.

Mayfield Creek's actual named trunk is short (~5 km, from where its forks join down to Loon Creek) and 100% Idaho Batholith granite -- no internal contact, so the automatic analysis doesn't recommend a trunk site. But the volcanic/granite complexity visible on the map is real and is happening on **East Fork Mayfield Creek and West Fork Mayfield Creek** (the grey context lines), each of which crosses once between the two units -- and both forks support spawning, so that variation is worth capturing rather than skipping.

Three manual sites now cover this drainage: one just below the fork confluence (on the short main-stem trunk), one on **West Fork Mayfield Creek**, and one on **East Fork Mayfield Creek**. The West Fork site sits a bit upstream of the volcanic/granite contact itself -- nudged there per Russ, to sit solidly inside the volcanic unit rather than right on the boundary, since the trunk and East Fork sites already characterize the plutonic side. It's still close to a road (the road-access reasoning that originally favored West Fork over East Fork here hasn't changed). The East Fork site is a new addition since Russ plans to hike up that fork a ways regardless -- worth a stop while he's up there even without it being tied to a specific contact.

## Camas Creek and sub-tributaries: WF Camas & Yellowjacket creeks

### Camas Creek

```{r}
#| label: fig-map-camas
map_drainage(drainage_defs[[8]])
```

```{r}
#| label: fig-camas
#| results: asis
report_drainage(drainage_defs[[8]])
```

The two auto-picked contacts (15, 38 km upstream) sit well above the Yellowjacket Creek confluence (~6.4 km) -- everything below that confluence is uniform Plutonic, so the added manual site there is for spatial coverage, not a lithologic contact. Russ may hike the whole drainage down to the river this September, and given how much variation both Camas and Yellowjacket show overall, it's worth grabbing a sample from the lower reach while he's passing through rather than leaving a several-km gap below the confluence unsampled.

### West Fork Camas Creek

```{r}
#| label: fig-map-wfcamas
map_drainage(drainage_defs[[9]])
```

```{r}
#| label: fig-wfcamas
#| results: asis
report_drainage(drainage_defs[[9]])
```

### Yellowjacket Creek

```{r}
#| label: fig-map-yellowjacket
map_drainage(drainage_defs[[10]])
```

```{r}
#| label: fig-yellowjacket
#| results: asis
report_drainage(drainage_defs[[10]])
```

West Fork Camas Creek's mapped trunk is 100% Challis volcanic rock -- geologically uniform like Rapid River, so no new site is recommended on lithologic grounds; place one by access if you want a point out there at all. (South Fork Camas Creek, which shares the "Camas Creek fork" naming pattern, was dropped from this plan entirely -- no salmon spawning there.)

Yellowjacket Creek stands out here: it's the only drainage in this year's target list with a substantial share (\~46%) of Precambrian Belt Supergroup metamorphic rock alongside Idaho Batholith granite. Per `docs/review_biogeochem_tracers.qmd`, ancient metamorphic terrain tends to be a highly radiogenic "anchor" — this is likely to be one of the most distinctive tributary signatures in the whole project, and worth prioritizing if time runs short.

## Wilson Creek

```{r}
#| label: fig-map-wilson
map_drainage(drainage_defs[[11]])
```

```{r}
#| label: fig-wilson
#| results: asis
report_drainage(drainage_defs[[11]])
```

Wilson Creek is short but geologically mixed (granite, metamorphic, and valley-fill all present) — a compact drainage where 2 well-placed sites should still capture most of the real variation. Lowest priority of Bryan's own list given how few spawners use it, but worth a stop if time allows given how much lithologic variation is packed into such a short reach.

## Marble Creek

```{r}
#| label: fig-map-marble
map_drainage(drainage_defs[[16]])
```

```{r}
#| label: fig-marble
#| results: asis
report_drainage(drainage_defs[[16]])
```

The trailhead site (top of drainage, 44.95702, -115.09996) is worth having regardless of who ends up collecting it -- reachable by Payette NF if they're already up there for Monumental Creek, or by Bryan otherwise. On the second question: yes, worth adding the site 7.4 km upstream of `MRB-010`. It isn't just an arbitrary "~5 km" spacing choice -- there's a real second volcanic/granite contact there, downstream of the existing 10.9 km candidate and past the small intervening volcanic patch you noticed. That gives three characterized contacts along Marble Creek instead of one.

## Big Creek and sub-tributaries

Note: "Rush Creek" and "Beaver Creek" are both common stream names in this basin -- the Big Creek versions used below are `GNIS_ID` 389731 and 377253 respectively (see the tributary identification reference near the top for the other streams sharing these names).

### Big Creek

```{r}
#| label: fig-map-big
map_drainage(drainage_defs[[12]])
```

```{r}
#| label: fig-big
#| results: asis
report_drainage(drainage_defs[[12]])
```

Big Creek itself is the most geologically diverse single tributary in this whole plan (five different rock categories along its length, including a meaningful metamorphic/sedimentary component) — this is exactly the kind of drainage where Baxter's crew collecting good spatial coverage along the mainstem will be valuable.

The blue triangles on the map are Hamann & Kennedy's (2012) own Big Creek water samples, added for reference against our candidates: our own 20.5 km candidate (45.13023, -114.9512) lines up reasonably well with their **Big Cr. upstream of Rush Cr. (site D)** (45.10520, -114.86843), and our 45 km candidate (45.1664, -115.2100) is a similarly good match to their **Big Creek upstream of Big Ramey (site C)** (45.17432, -115.18269) — good agreement between independently-derived candidate spots and a decade-old published dataset, and a useful cross-check on the whole approach. Since D and C are effectively already covered, they aren't added as separate sites; instead, three of Hamann & Kennedy's other sites that land in gaps our own contact analysis didn't fill are added directly to the candidate list: **Lower Big Cr. (site E)** (45.10261, -114.84607, 8.7 km upstream), **Upper Big Cr. (site A)** (45.10630, -115.32839, 60.9 km upstream), and **site B**, downstream of Logan Cr. (45.12779, -115.31390, 58.2 km upstream). A sixth site, ~30.9 km upstream (45.14177, -115.05557), fills the largest remaining gap -- it sits within the volcanic stretch between the 20.5 km and 45 km contact-based candidates, so it's spatial coverage rather than a new contact. That brings Big Creek to 6 candidate sites total (3 auto-picked/spatial + 3 from Hamann & Kennedy), alongside the existing `BIG-010` confluence site — good spatial coverage for Baxter's crew across the full length of the mainstem.

### Rush Creek

```{r}
#| label: fig-map-rush
map_drainage(drainage_defs[[13]])
```

```{r}
#| label: fig-rush
#| results: asis
report_drainage(drainage_defs[[13]])
```

Rush Creek itself is remote and hard to reach beyond its lower end. Hamann & Kennedy's "Rush Cr." reference site (45.08826, -114.88196) sits right at the Big Creek confluence and is reachable via the ranch/trail network, so it's added to the candidate list as the **priority stop** on this drainage. The two contact-based candidates further upstream (6.2, 14.2 km) are geologically meaningful (Metasedimentary/Plutonic and Volcanic/Metasedimentary contacts) but secondary — worth pushing for if access allows, but not expected to be reachable most seasons.

### Beaver Creek (Big Ck drainage)

```{r}
#| label: fig-map-beaverbig
map_drainage(drainage_defs[[14]])
```

```{r}
#| label: fig-beaverbig
#| results: asis
report_drainage(drainage_defs[[14]])
```

Same situation as Rush Creek: Hamann & Kennedy's "Beaver Cr." reference site (45.17340, -115.24476) sits right at the Big Creek confluence, so it's added as the priority candidate on this Beaver Creek (`BVB` — not to be confused with Marsh Creek's `BVR`, see `data/site_id_convention.md`). The two contact-based candidates further upstream (7.9, 13.7 km) are secondary, same access logic as Rush Creek.

### Monumental Creek

```{r}
#| label: fig-map-monumental
map_drainage(drainage_defs[[15]])
```

```{r}
#| label: fig-monumental
#| results: asis
report_drainage(drainage_defs[[15]])
```

Three tiers of priority here. **Lower** (7.9 km upstream): Hamann & Kennedy's "Monumental Cr. (site F)" reference site (45.09937, -115.15733) is now added to the candidate list — it's reachable from the Big Creek side, so this is doable even without going deep into the Monumental Creek drainage itself. **Upper** (24.5 km upstream): the manual site near the Monumental/Marble trailhead (44.98076, -115.16526) is reachable via the trailhead. **Middle** (14.9 km upstream, the analysis-picked Volcanic/Metasedimentary contact): geologically the most interesting of the three, but sitting well inside the drainage between the other two, it's likely the hardest to reach this season — worth attempting only if time and access allow after the lower and upper sites are covered.

## Punch list

```{r}
#| label: tbl-punchlist
punchlist <- map_dfr(drainage_defs, function(def) {
  r <- results[[def$drainage]]
  cand <- get_candidates(def)
  tibble(
    drainage = def$drainage,
    code = def$code,
    mapped_km = if (!is.null(r)) r$total_len_km else NA,
    dominant_unit = if (!is.null(r)) r$composition$lith_group[1] else NA,
    n_contacts = if (!is.null(r)) nrow(r$crossings) else NA,
    n_candidate_sites = nrow(cand),
    partner = if (!is.null(def$partner)) def$partner else "Bryan (own itinerary)"
  )
})
kable(punchlist)
```

Priorities if the season runs short on time: **Yellowjacket Creek** (only metamorphic-heavy drainage in Bryan's own list) and **Indian Creek** (cleanest single volcanic/granite contact) give the most new isotopic information per site visited. **Rapid River** is the lowest-priority stop for a *new* upstream site specifically, since it's geologically uniform — the existing `RAP-010` already characterizes it well.

## All candidate sites at a glance

Same view as the overview map above, with every candidate new site from every section overlaid -- a single quick-reference for the whole season rather than paging through each drainage individually.

```{r}
#| label: fig-map-all-candidates
#| fig-width: 9
#| fig-height: 9
#| message: false

all_candidates <- map_dfr(drainage_defs, function(def) {
  cand <- get_candidates(def)
  if (nrow(cand) == 0) return(NULL)
  cand |> filter(!is.na(lat)) |> mutate(drainage = def$drainage, code = def$code)
})
all_candidates_sf <- all_candidates |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(st_crs(fl))

ggplot() +
  geom_sf(data = lith_clip, aes(fill = lith_group), color = NA, alpha = 0.6) +
  geom_sf(data = hu8, fill = NA, color = "grey30", linewidth = 0.5, linetype = "dashed") +
  geom_sf(data = fl, color = "grey20", linewidth = 0.15) +
  geom_sf(data = fl |> filter(GNIS_NAME == "Middle Fork Salmon River"), color = "blue", linewidth = 1) +
  geom_sf(data = target_segs, color = "blue", linewidth = 0.5) +
  geom_sf(data = existing_sites_sf, color = "black", size = 1.5) +
  geom_sf(data = all_candidates_sf, color = "red", size = 2, shape = 4, stroke = 1.3) +
  scale_fill_manual(values = litho_colors, name = "Bedrock unit") +
  map_furniture() +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right", legend.text = element_text(size = 7),
    plot.caption = element_text(hjust = 0, size = 7)
  ) +
  guides(fill = guide_legend(nrow = 5)) +
  labs(
    title = "MFSR basin: all candidate new sites, 2026 season",
    caption = str_wrap("Black dots = existing sites. Red x's = candidate new sites (this doc). Blue lines = target tributaries + named forks.", width = 100)
  )
```

## 2026 site locations: table and KML

Every existing established site plus every candidate new site from this doc, in one place -- a single reference table for planning, and a KML file below that loads directly into OnX, Google Earth, or similar field mapping software.

```{r}
#| label: tbl-all-sites-2026
#| message: false

existing_table <- sites |>
  transmute(
    Type = "Existing",
    Drainage = drainage_group,
    `Site ID` = site_id,
    lat = lat_dd, lon = long_dd,
    Notes = paste0(waterbody_name, " -- established ", date_established)
  )

candidate_table <- all_candidates |>
  arrange(drainage, dist_upstream_km) |>
  mutate(seq = row_number(), .by = code) |>
  transmute(
    Type = "Candidate (2026)",
    Drainage = drainage,
    `Site ID` = paste0(code, "-NEW", seq),
    lat, lon,
    Notes = rationale
  )

all_sites_table <- bind_rows(existing_table, candidate_table) |> arrange(Drainage, Type)

all_sites_table |>
  select(Type, Drainage, `Site ID`, lat, lon, Notes) |>
  kable(digits = 5)
```

```{r}
#| label: export-kml
#| message: false
#| results: hide

kml_sites <- all_sites_table |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  transmute(Name = `Site ID`, Description = paste0(Drainage, " (", Type, "): ", Notes))

kml_path <- here("data/spatial/mfsr_2026_sites.kml")
if (file.exists(kml_path)) file.remove(kml_path)
st_write(kml_sites, kml_path, driver = "KML", quiet = TRUE)
```

Download: [`data/spatial/mfsr_2026_sites.kml`](../data/spatial/mfsr_2026_sites.kml) -- regenerated automatically every time this doc renders, so it always matches the table and maps above. Existing sites use their real `site_id`; candidate sites use a placeholder ID (`[CODE]-NEW1`, `[CODE]-NEW2`, ...) following the same convention as `field/partner_sample_collection_guide.qmd` -- replace with the real assigned ID once a site is actually sampled.
